---
title: "Hands-on Exercise 2A: First-order Spatial Point Patterns Analysis"
author: "Bao Sihan"
date: "September 1, 2026"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Overview

In this hands-on exercise, I examine the first-order spatial pattern of childcare centres in Singapore. First-order spatial point pattern analysis focuses on how the intensity of events varies across space. I use the Clark-Evans test to examine whether the locations are randomly distributed and kernel density estimation (KDE) to identify areas with higher concentrations of childcare centres.

The analysis follows two questions:

- Are childcare centres randomly distributed across Singapore?
- If they are not randomly distributed, where are the areas with higher concentrations?

## The Data

I use the two data sets specified in the exercise:

- **Child Care Services** from [data.gov.sg](https://data.gov.sg/datasets/d_5d668e3f544335f8028f546827b773b4/view), which provides the locations and attributes of childcare centres; and
- **Master Plan 2019 Subzone Boundary (No Sea)** from [data.gov.sg](https://data.gov.sg/datasets/d_8594ae9ff96d0c708bc2af633048edfb/view), which provides Singapore's planning subzone boundaries.

Both files are stored in GeoJSON format inside this exercise's `data/geospatial` folder.

## Installing and Loading the R Packages

I use **sf** to manage vector data, **spatstat** for spatial point pattern analysis, **terra** to convert the KDE image to a raster, **tmap** for mapping, **rvest** as listed in the exercise, and **tidyverse** for data wrangling.

```{r}
#| label: load-packages
pacman::p_load(
  sf,
  terra,
  spatstat,
  tmap,
  rvest,
  tidyverse
)
```

## Importing and Wrangling Geospatial Data Sets

### Importing the Master Plan 2019 subzone boundaries

I import the subzone boundaries with `st_read()` and transform them from WGS 84 to Singapore SVY21 (EPSG:3414). The GeoJSON file already stores the required planning attributes in separate columns, so no KML description fields need to be extracted.

```{r}
#| label: import-subzones
mpsz_sf <- st_read(
  "data/geospatial/MasterPlan2019SubzoneBoundaryNoSea.geojson"
) %>%
  st_transform(crs = 3414) %>%
  st_make_valid() %>%
  select(
    REGION_N,
    PLN_AREA_N,
    SUBZONE_N,
    SUBZONE_C,
    geometry
  )

mpsz_sf
```

I remove the offshore planning areas used in the professor's workflow before creating the Singapore observation window.

```{r}
#| label: clean-subzones
mpsz_cl <- mpsz_sf %>%
  filter(
    SUBZONE_N != "SOUTHERN GROUP",
    PLN_AREA_N != "WESTERN ISLANDS",
    PLN_AREA_N != "NORTH-EASTERN ISLANDS"
  )

dir.create("data/rds", recursive = TRUE, showWarnings = FALSE)
write_rds(mpsz_cl, "data/rds/mpsz_cl.rds")
```

### Importing the childcare locations

I import the childcare centre locations and transform them to the same projected coordinate reference system.

```{r}
#| label: import-childcare
childcare_sf <- st_read(
  "data/geospatial/ChildCareServices.geojson"
) %>%
  st_transform(crs = 3414)

childcare_sf
```

The two layers now have the same CRS.

```{r}
#| label: check-crs
st_crs(mpsz_cl)$epsg
st_crs(childcare_sf)$epsg
```

### Mapping the geospatial data sets

I first prepare a static map to confirm that the childcare locations fall within the same spatial extent as the subzone boundaries.

```{r}
#| label: map-input-data
#| fig-width: 8
#| fig-height: 7
tmap_mode("plot")

tm_shape(mpsz_cl) +
  tm_polygons(fill = NA, col = "grey50", lwd = 0.5) +
  tm_shape(childcare_sf) +
  tm_dots(fill = "red", size = 0.02) +
  tm_layout(frame = FALSE)
```

The childcare points and subzone boundaries occupy the same spatial context, confirming that the projection and coordinate values are compatible.

I also display the childcare locations in interactive mode and then return `tmap` to plot mode.

```{r}
#| label: interactive-childcare-map
tmap_mode("view")

tm_shape(childcare_sf) +
  tm_dots()

tmap_mode("plot")
```

## Geospatial Data Wrangling for `spatstat`

### Converting the point data to `ppp`

`spatstat` represents planar point patterns with the `ppp` class. I convert the childcare simple feature data frame and inspect the resulting object.

```{r}
#| label: convert-childcare-ppp
childcare_ppp <- as.ppp(childcare_sf)

class(childcare_ppp)
summary(childcare_ppp)
```

### Creating the observation window

I convert the cleaned Singapore subzone boundaries into an `owin` object. This confines the analysis to the land area represented by the selected subzones.

```{r}
#| label: create-singapore-window
sg_owin <- as.owin(mpsz_cl)

class(sg_owin)
```

```{r}
#| label: plot-singapore-window
#| fig-width: 7
#| fig-height: 7
plot(sg_owin, main = "Singapore observation window")
```

### Combining the point pattern and observation window

I retain the childcare events that fall within the Singapore observation window.

```{r}
#| label: clip-childcare-points
childcareSG_ppp <- childcare_ppp[sg_owin]

childcareSG_ppp
```

## Clark-Evans Test for Nearest Neighbour Analysis

The Clark-Evans test compares the observed mean nearest-neighbour distance with the distance expected under complete spatial randomness. I use the following hypotheses:

- $H_0$: Childcare centres are randomly distributed.
- $H_1$: Childcare centres are spatially clustered.

I use a 95% confidence level.

### Clark-Evans test without CSR simulation

```{r}
#| label: clark-evans-no-csr
clark_no_csr <- clarkevans.test(
  childcareSG_ppp,
  correction = "none",
  clipregion = sg_owin,
  alternative = "clustered"
)

clark_no_csr
```

The Clark-Evans ratio is `r round(unname(clark_no_csr$statistic), 3)` and the p-value is below 0.05. I therefore reject the null hypothesis. The ratio is also below 1, which indicates that childcare centres are clustered rather than randomly distributed.

From a planning perspective, this means that childcare provision is concentrated in some parts of Singapore. Areas outside these concentrations may need to be reviewed when considering the accessibility of childcare services.

### Clark-Evans test with CSR simulation

I repeat the test using 99 Monte Carlo simulations of complete spatial randomness.

```{r}
#| label: clark-evans-csr
set.seed(1234)

clark_csr <- clarkevans.test(
  childcareSG_ppp,
  correction = "none",
  clipregion = sg_owin,
  alternative = "clustered",
  method = "MonteCarlo",
  nsim = 99
)

clark_csr
```

The simulated test gives a p-value of `r format.pval(clark_csr$p.value, digits = 3)`, which is below 0.05. This supports the same conclusion: the childcare centre pattern is significantly clustered.

## Kernel Density Estimation

KDE converts the childcare points into a continuous surface so that areas with higher point intensity can be identified.

### Automatic bandwidth selection

I first use `bw.diggle` to select the bandwidth automatically. A Gaussian kernel and edge correction are applied.

```{r}
#| label: kde-diggle-metres
#| fig-width: 8
#| fig-height: 6
kde_SG_diggle <- density(
  childcareSG_ppp,
  sigma = bw.diggle,
  edge = TRUE,
  kernel = "gaussian"
)

plot(kde_SG_diggle, main = "Childcare centre KDE using bw.diggle")
```

The result is an `im` object containing a two-dimensional grid of density values.

```{r}
#| label: inspect-kde
summary(kde_SG_diggle)

bw <- bw.diggle(childcareSG_ppp)
bw
```

The density values are initially expressed as points per square metre, so the numbers are very small.

### Rescaling the KDE values

I rescale the point pattern from metres to kilometres and recompute the KDE. The resulting density is easier to interpret as the number of childcare centres per square kilometre.

```{r}
#| label: rescale-to-kilometres
childcareSG_ppp_km <- rescale.ppp(
  childcareSG_ppp,
  1000,
  "km"
)

kde_childcareSG_km <- density(
  childcareSG_ppp_km,
  sigma = bw.diggle,
  edge = TRUE,
  kernel = "gaussian"
)
```

```{r}
#| label: plot-kde-kilometres
#| fig-width: 8
#| fig-height: 6
plot(
  kde_childcareSG_km,
  main = "Childcare centre density per square kilometre"
)
```

### Comparing automatic bandwidth methods

I compare the four bandwidth selectors introduced in the exercise.

```{r}
#| label: compare-bandwidth-values
bw.CvL(childcareSG_ppp_km)
bw.scott(childcareSG_ppp_km)
bw.ppl(childcareSG_ppp_km)
bw.diggle(childcareSG_ppp_km)
```

`bw.ppl()` generally gives a useful bandwidth when a point pattern contains tight clusters, while `bw.diggle()` is useful for detecting a tight cluster among background noise. I compare their KDE surfaces below.

```{r}
#| label: compare-diggle-ppl
#| fig-width: 10
#| fig-height: 5
kde_childcareSG_ppl <- density(
  childcareSG_ppp_km,
  sigma = bw.ppl,
  edge = TRUE,
  kernel = "gaussian"
)

par(mfrow = c(1, 2))
plot(kde_childcareSG_km, main = "bw.diggle")
plot(kde_childcareSG_ppl, main = "bw.ppl")
par(mfrow = c(1, 1))
```

The two bandwidths produce different levels of smoothing. The `bw.diggle` result retains more local variation, while the `bw.ppl` surface is smoother.

### Comparing kernel functions

Keeping the bandwidth at the value used in the professor's example, I compare Gaussian, Epanechnikov, quartic, and disc kernels.

```{r}
#| label: compare-kernel-functions
#| fig-width: 10
#| fig-height: 9
par(mfrow = c(2, 2))

plot(
  density(
    childcareSG_ppp_km,
    sigma = 0.2959712,
    edge = TRUE,
    kernel = "gaussian"
  ),
  main = "Gaussian"
)

plot(
  density(
    childcareSG_ppp_km,
    sigma = 0.2959712,
    edge = TRUE,
    kernel = "epanechnikov"
  ),
  main = "Epanechnikov"
)

plot(
  density(
    childcareSG_ppp_km,
    sigma = 0.2959712,
    edge = TRUE,
    kernel = "quartic"
  ),
  main = "Quartic"
)

plot(
  density(
    childcareSG_ppp_km,
    sigma = 0.2959712,
    edge = TRUE,
    kernel = "disc"
  ),
  main = "Disc"
)

par(mfrow = c(1, 1))
```

The overall high- and low-density areas remain similar, although the shape and smoothness of the surfaces vary with the kernel function.

## Fixed and Adaptive KDE

### Fixed bandwidth KDE

I use a fixed bandwidth of 0.6 kilometres, which is equivalent to 600 metres.

```{r}
#| label: fixed-bandwidth-kde
#| fig-width: 8
#| fig-height: 6
kde_childcareSG_fb <- density(
  childcareSG_ppp_km,
  sigma = 0.6,
  edge = TRUE,
  kernel = "gaussian"
)

plot(kde_childcareSG_fb, main = "Fixed bandwidth KDE")
```

### Adaptive bandwidth KDE

An adaptive bandwidth changes according to the local point distribution. This is useful when the pattern contains areas with very different point densities.

```{r}
#| label: adaptive-bandwidth-kde
#| fig-width: 8
#| fig-height: 6
kde_childcareSG_ab <- adaptive.density(
  childcareSG_ppp_km,
  method = "kernel"
)

plot(kde_childcareSG_ab, main = "Adaptive bandwidth KDE")
```

I compare the fixed and adaptive results directly.

```{r}
#| label: compare-fixed-adaptive
#| fig-width: 10
#| fig-height: 5
par(mfrow = c(1, 2))
plot(kde_childcareSG_fb, main = "Fixed bandwidth")
plot(kde_childcareSG_ab, main = "Adaptive bandwidth")
par(mfrow = c(1, 1))
```

The fixed bandwidth applies the same amount of smoothing throughout Singapore, while the adaptive result adjusts the smoothing to local variations in the childcare point pattern.

## Plotting a Cartographic KDE Map

### Converting the gridded output to a raster

I convert the rescaled KDE image into a `SpatRaster` object.

```{r}
#| label: convert-kde-raster
kde_childcareSG_bw_terra <- rast(kde_childcareSG_km)

class(kde_childcareSG_bw_terra)
kde_childcareSG_bw_terra
```

### Assigning the projection

I assign Singapore SVY21 to the raster, following the exercise workflow.

```{r}
#| label: assign-raster-crs
crs(kde_childcareSG_bw_terra) <- "EPSG:3414"

kde_childcareSG_bw_terra
```

### Mapping the KDE raster with `tmap`

```{r}
#| label: map-kde-raster
#| fig-width: 9
#| fig-height: 7
tm_shape(kde_childcareSG_bw_terra) +
  tm_raster(
    col.scale = tm_scale_continuous(values = "viridis"),
    col.legend = tm_legend(
      title = "Density values",
      title.size = 0.7,
      text.size = 0.7,
      bg.color = "white",
      bg.alpha = 0.7,
      position = tm_pos_in("right", "bottom"),
      frame = TRUE
    )
  ) +
  tm_graticules(labels.size = 0.7) +
  tm_compass() +
  tm_layout(scale = 1)
```

The cartographic map makes the variation in childcare centre density easier to compare across the study area.

## First-order SPPA at the Planning Area Level

### Extracting the study areas

I focus on Punggol, Tampines, Choa Chu Kang, and Jurong West, as specified in the exercise.

```{r}
#| label: extract-planning-areas
pg <- mpsz_cl %>% filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_cl %>% filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl %>% filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_cl %>% filter(PLN_AREA_N == "JURONG WEST")
```

```{r}
#| label: plot-planning-areas
#| fig-width: 10
#| fig-height: 9
par(mfrow = c(2, 2))
plot(st_geometry(pg), main = "Punggol")
plot(st_geometry(tm), main = "Tampines")
plot(st_geometry(ck), main = "Choa Chu Kang")
plot(st_geometry(jw), main = "Jurong West")
par(mfrow = c(1, 1))
```

### Creating planning-area observation windows

```{r}
#| label: create-planning-windows
pg_owin <- as.owin(pg)
tm_owin <- as.owin(tm)
ck_owin <- as.owin(ck)
jw_owin <- as.owin(jw)
```

### Combining the points and observation windows

I clip the national childcare point pattern to each planning area and rescale the four point patterns to kilometres for plotting and KDE.

```{r}
#| label: clip-planning-points
childcare_pg_ppp <- childcare_ppp[pg_owin]
childcare_tm_ppp <- childcare_ppp[tm_owin]
childcare_ck_ppp <- childcare_ppp[ck_owin]
childcare_jw_ppp <- childcare_ppp[jw_owin]

childcare_pg_ppp_km <- rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp_km <- rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp_km <- rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp_km <- rescale.ppp(childcare_jw_ppp, 1000, "km")
```

```{r}
#| label: plot-planning-points
#| fig-width: 10
#| fig-height: 9
par(mfrow = c(2, 2))
plot(unmark(childcare_pg_ppp_km), main = "Punggol")
plot(unmark(childcare_tm_ppp_km), main = "Tampines")
plot(unmark(childcare_ck_ppp_km), main = "Choa Chu Kang")
plot(unmark(childcare_jw_ppp_km), main = "Jurong West")
par(mfrow = c(1, 1))
```

### Clark-Evans test for Choa Chu Kang

```{r}
#| label: clark-evans-choa-chu-kang
clark_ck <- clarkevans.test(
  childcare_ck_ppp,
  correction = "none",
  clipregion = NULL,
  alternative = "two.sided",
  nsim = 999
)

clark_ck
```

The Clark-Evans ratio for Choa Chu Kang is `r round(unname(clark_ck$statistic), 3)` and its p-value is below 0.05. I reject the null hypothesis of spatial randomness. Since the ratio is below 1, the childcare centres show a clustered pattern.

### Clark-Evans test for Tampines

```{r}
#| label: clark-evans-tampines
clark_tm <- clarkevans.test(
  childcare_tm_ppp,
  correction = "none",
  clipregion = NULL,
  alternative = "two.sided",
  nsim = 999
)

clark_tm
```

The Tampines ratio is `r round(unname(clark_tm$statistic), 3)` and its p-value is also below 0.05. I reject the null hypothesis and conclude that childcare centres in Tampines are clustered.

### Computing KDE surfaces by planning area

Finally, I use `bw.diggle` with a Gaussian kernel to compare the childcare density surfaces for the four planning areas.

```{r}
#| label: planning-area-kde
#| fig-width: 10
#| fig-height: 9
par(mfrow = c(2, 2))

plot(
  density(
    childcare_pg_ppp_km,
    sigma = bw.diggle,
    edge = TRUE,
    kernel = "gaussian"
  ),
  main = "Punggol"
)

plot(
  density(
    childcare_tm_ppp_km,
    sigma = bw.diggle,
    edge = TRUE,
    kernel = "gaussian"
  ),
  main = "Tampines"
)

plot(
  density(
    childcare_ck_ppp_km,
    sigma = bw.diggle,
    edge = TRUE,
    kernel = "gaussian"
  ),
  main = "Choa Chu Kang"
)

plot(
  density(
    childcare_jw_ppp_km,
    sigma = bw.diggle,
    edge = TRUE,
    kernel = "gaussian"
  ),
  main = "Jurong West"
)

par(mfrow = c(1, 1))
```

The four KDE surfaces show that the intensity of childcare centres varies within each planning area rather than remaining spatially uniform.
