---
title: "Hands-on Exercise 4B: Visualising Distribution"
author: "Bao Sihan"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
execute:
  warning: false
  message: false
---

# 9.1 Overview

In this hands-on exercise, I practise two distribution visualisation methods using `Exam_data.csv`:

-   **Ridgeline plots** (ggridges)
-   **Raincloud plots** (ggdist)

# 9.2 Installing and launching R packages

```{r}
pacman::p_load(ggdist, ggridges, ggthemes, colorspace, tidyverse)
```

# 9.3 Importing data

```{r}
exam <- read_csv("../data/Exam_data.csv")
glimpse(exam)
```

# 9.4 Ridgeline plots

## 9.4.1 Basic ridgeline plot

```{r}
ggplot(exam, aes(x = ENGLISH, y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", 0.3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

## 9.4.2 Varying fill colors along the x axis

```{r}
ggplot(exam, aes(x = ENGLISH, y = CLASS, fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01
  ) +
  scale_fill_viridis_c(name = "Temp. [F]", option = "C") +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

## 9.4.3 Mapping tail probabilities (ecdf)

```{r}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = 0.5 - abs(0.5 - stat(ecdf)))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE
  ) +
  scale_fill_viridis_c(
    name = "Tail probability",
    direction = -1
  ) +
  theme_ridges()
```

## 9.4.4 Ridgeline plots with quantile lines

```{r}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = factor(stat(quantile)))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = 4,
    quantile_lines = TRUE
  ) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```

# 9.5 Raincloud plots

## 9.5.1 Half-eye

```{r}
ggplot(exam, aes(x = RACE, y = ENGLISH)) +
  stat_halfeye(
    adjust = 0.5,
    justification = -0.2,
    .width = 0,
    point_colour = NA
  )
```

## 9.5.2 Add boxplot

```{r}
ggplot(exam, aes(x = RACE, y = ENGLISH)) +
  stat_halfeye(
    adjust = 0.5,
    justification = -0.2,
    .width = 0,
    point_colour = NA
  ) +
  geom_boxplot(width = 0.20, outlier.shape = NA)
```

## 9.5.3 Add dot plot

```{r}
ggplot(exam, aes(x = RACE, y = ENGLISH)) +
  stat_halfeye(
    adjust = 0.5,
    justification = -0.2,
    .width = 0,
    point_colour = NA
  ) +
  geom_boxplot(width = 0.20, outlier.shape = NA) +
  stat_dots(
    side = "left",
    justification = 1.2,
    binwidth = 0.5,
    dotsize = 1.5
  )
```

## 9.5.4 Finishing touch

```{r}
ggplot(exam, aes(x = RACE, y = ENGLISH)) +
  stat_halfeye(
    adjust = 0.5,
    justification = -0.2,
    .width = 0,
    point_colour = NA
  ) +
  geom_boxplot(width = 0.20, outlier.shape = NA) +
  stat_dots(
    side = "left",
    justification = 1.2,
    binwidth = 0.5,
    dotsize = 1.5
  ) +
  coord_flip() +
  theme_economist()
```
