---
title: "Hands-on Exercise 4: Spatial Weights and Applications"
author: "Bao Sihan"
date: "September 17, 2026"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
  fig-height: 6
---

## Overview

In this exercise, I use Hunan's county boundaries and 2012 development indicators to explore how spatial weights describe relationships between areas. I first define neighbours using shared boundaries and distances. I then use these relationships to calculate spatially lagged GDP per capita (GDPPC), including neighbourhood averages and sums.

The exercise follows [Chapter 8: Spatial Weights and Applications](https://r4gdsa.netlify.app/chap08.html). My focus is on understanding what each weight and lag represents.

## The Study Area and Data

I use the two data sets provided for the exercise:

- **Hunan county boundaries**, an ESRI shapefile containing 88 county polygons.
- **Hunan_2012.csv**, a table of 2012 development indicators for the same 88 counties.

The shapefile and its companion files are in `data/geospatial`, while the CSV and the supplied data dictionary are in `data/aspatial`. All file paths below are relative to this exercise folder.

### Getting started

I load **sf** for spatial data, **spdep** for neighbour lists and spatial weights, **tmap** for maps, **tidyverse** for data preparation, and **knitr** for tables.

```{r}
#| label: load-packages
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
tmap_mode("plot")
```

## Getting the Data into R

### Importing the shapefile

```{r}
#| label: import-boundaries
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan")
hunan
```

### Importing the development indicators

```{r}
#| label: import-indicators
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
glimpse(hunan2012)
```

### Joining the data

I join the two data sets by `County` and keep the fields used in the exercise. Naming these fields makes it easier to see which attributes are retained.

```{r}
#| label: join-data
hunan <- left_join(hunan, hunan2012, by = "County") %>%
  select(NAME_2, ID_3, NAME_3, ENGTYPE_3, County, GDPPC)
hunan
```

```{r}
#| label: check-data
#| include: false
stopifnot(
  nrow(hunan) == 88L,
  nrow(hunan2012) == 88L,
  !anyDuplicated(hunan$County),
  !anyDuplicated(hunan2012$County),
  !anyNA(hunan$GDPPC),
  all(st_is_valid(hunan)),
  st_is_longlat(hunan)
)
```

The joined layer contains `r nrow(hunan)` counties, with no missing GDPPC values.

## Visualising the Regional Development Indicator

I compare a labelled county map with a choropleth of GDPPC. The labels help me locate the counties before examining their neighbour relationships. I use the current **tmap** syntax for the legend position.

```{r}
#| label: regional-development-maps
#| fig-cap: "Hunan county names and GDP per capita in 2012."
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size = 0.5)

gdppc <- qtm(hunan, fill = "GDPPC") +
  tm_layout(legend.position = tm_pos_in("left", "bottom"))

tmap_arrange(basemap, gdppc, asp = 1, ncol = 2)
```

## Computing Contiguity Spatial Weights

### Queen contiguity

Queen contiguity treats counties as neighbours when their boundaries share either an edge or a vertex.

```{r}
#| label: queen-neighbours
wm_q <- poly2nb(hunan, queen = TRUE)
summary(wm_q)
```

There are `r length(wm_q)` counties and `r sum(card(wm_q))` directed neighbour links. The number of neighbours ranges from `r min(card(wm_q))` to `r max(card(wm_q))`.

I examine the first county and its neighbours in more detail.

```{r}
#| label: first-county-neighbours
wm_q[[1]]
hunan$County[1]
hunan$NAME_3[wm_q[[1]]]

nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

The first county is **`r hunan$County[1]`**. The integers in `wm_q[[1]]` refer to row positions in `hunan`, so I use them to retrieve the neighbouring county names and GDPPC values.

```{r}
#| label: queen-list-structure
str(wm_q)
```

### Rook contiguity

Rook contiguity requires a shared boundary edge. Counties that meet only at a vertex are not considered neighbours under this definition.

```{r}
#| label: rook-neighbours
wm_r <- poly2nb(hunan, queen = FALSE)
summary(wm_r)
```

The Rook list contains `r sum(card(wm_r))` directed links, compared with `r sum(card(wm_q))` under Queen contiguity. Its largest neighbour count is `r max(card(wm_r))`.

### Visualising the contiguity weights

To draw the links, I calculate one centroid for each county, following the coordinate extraction used in the exercise. These longitude and latitude coordinates provide the endpoints for the connectivity graphs.

```{r}
#| label: county-centroids
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
coords <- cbind(longitude, latitude)
head(coords)
```

#### Queen neighbours

```{r}
#| label: queen-connectivity-map
#| fig-cap: "Connections between Queen-contiguous counties."
plot(hunan$geometry, border = "lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

#### Rook neighbours

```{r}
#| label: rook-connectivity-map
#| fig-cap: "Connections between Rook-contiguous counties."
plot(hunan$geometry, border = "lightgrey")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

#### Comparing Queen and Rook neighbours

```{r}
#| label: contiguity-comparison
#| fig-cap: "Queen and Rook connectivity graphs, shown side by side."
par(mfrow = c(1, 2))
plot(hunan$geometry, border = "lightgrey", main = "Queen contiguity")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
plot(hunan$geometry, border = "lightgrey", main = "Rook contiguity")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
par(mfrow = c(1, 1))
```

## Computing Distance-based Neighbours

### Determining the cut-off distance

I first identify each county's nearest neighbour, then measure those links in kilometres using `longlat = TRUE`. The largest nearest-neighbour distance helps determine a threshold that leaves no county without a neighbour.

```{r}
#| label: nearest-neighbour-distances
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

The largest distance is approximately **`r round(max(k1dists), 2)` km**. Following the exercise, I use a 62 km distance band.

### Fixed-distance weights

```{r}
#| label: fixed-distance-neighbours
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
str(wm_d62)
```

The average number of links is `r round(mean(card(wm_d62)), 6)`. This means a county has about `r round(mean(card(wm_d62)), 2)` other county centroids within 62 km on average; it does not mean every county has that number of neighbours.

```{r}
#| label: fixed-distance-counts
table(hunan$County, card(wm_d62))

n_comp <- n.comp.nb(wm_d62)
n_comp$nc
table(n_comp$comp.id)
```

Here, all 88 counties belong to one connected component. In general, giving every county at least one neighbour does not guarantee that the entire graph is connected, so I check both properties.

#### Plotting the fixed-distance links

```{r}
#| label: fixed-distance-overlay
#| fig-cap: "Links within 62 km (black), with first-nearest-neighbour links overlaid in red."
plot(hunan$geometry, border = "lightgrey")
plot(wm_d62, coords, add = TRUE)
plot(k1, coords, add = TRUE, col = "red", length = 0.08)
```

```{r}
#| label: distance-comparison
#| fig-cap: "First-nearest-neighbour links and the 62 km distance graph."
par(mfrow = c(1, 2))
plot(hunan$geometry, border = "lightgrey", main = "1st nearest neighbours")
plot(k1, coords, add = TRUE, col = "red", length = 0.08)
plot(hunan$geometry, border = "lightgrey", main = "Distance links: 62 km")
plot(wm_d62, coords, add = TRUE, pch = 19, cex = 0.6)
par(mfrow = c(1, 1))
```

### Adaptive-distance weights

A fixed distance can produce different numbers of neighbours across counties. The six-nearest-neighbour approach instead gives every county six neighbours, allowing the distance to the furthest neighbour to vary. These links need not be reciprocal.

```{r}
#| label: six-nearest-neighbours
knn6 <- knn2nb(knearneigh(coords, k = 6))
knn6
str(knn6)
```

#### Plotting the six-nearest-neighbour links

```{r}
#| label: adaptive-distance-map
#| fig-cap: "Each county is linked to its six nearest neighbours."
plot(hunan$geometry, border = "lightgrey")
plot(knn6, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

## Weights Based on Inverse Distance

Here I return to the Queen neighbour list and calculate the distance of each link. Taking the inverse of distance assigns a larger weight to closer neighbours. This changes the weights on existing Queen links, rather than adding new neighbours.

```{r}
#| label: inverse-distance-weights
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1 / x)
ids
```

## Row-standardised Weights

With `style = "W"`, equal weights are divided by the number of neighbours in each row. The weights therefore sum to one for every county that has neighbours.

```{r}
#| label: row-standardised-queen
rswm_q <- nb2listw(wm_q, style = "W", zero.policy = TRUE)
rswm_q
rswm_q$weights[10]
```

County 10, **`r hunan$County[10]`**, has `r card(wm_q)[10]` neighbours. Each receives a weight of `r signif(rswm_q$weights[[10]][1], 4)`. The `zero.policy = TRUE` option allows an area with no neighbours to receive a zero lag, although the Queen list here contains no such areas.

I also create the inverse-distance weights object shown in the exercise.

```{r}
#| label: inverse-distance-listw
rswm_ids <- nb2listw(wm_q, glist = ids, style = "B", zero.policy = TRUE)
rswm_ids
rswm_ids$weights[1]
summary(unlist(rswm_ids$weights))
```

Despite the object name, this second object is **not row-standardised**: `style = "B"` retains the supplied inverse-distance values without normalising each row. The `rswm_q` object is the row-standardised one used in the next calculation.

## Applications of the Spatial Weights Matrix

### Spatial lag with row-standardised weights

I calculate the weighted average GDPPC of each county's Queen neighbours with `lag.listw()`.

```{r}
#| label: lag-average
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
GDPPC.lag

nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

For `r hunan$County[1]`, the spatial lag is **`r format(round(GDPPC.lag[1], 1), big.mark = ",", trim = TRUE)`**. This is the arithmetic mean of its `r length(nb1)` neighbours' GDPPC values, because all neighbours receive equal row-standardised weights. The county's own value is excluded.

```{r}
#| label: join-lag-average
lag.list <- list(hunan$NAME_3, GDPPC.lag)
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan, lag.res, by = "NAME_3")
head(hunan)
```

```{r}
#| label: lag-average-maps
#| fig-cap: "County GDPPC and the average GDPPC of Queen neighbours. Each map has its own legend."
gdppc <- qtm(hunan, fill = "GDPPC")
lag_gdppc <- qtm(hunan, fill = "lag GDPPC")
tmap_arrange(gdppc, lag_gdppc, asp = 1, ncol = 2)
```

The lag map summarises the surroundings of each county. I read the legends separately because the original values and neighbouring averages have different ranges.

### Spatial lag as a sum of neighbouring values

To calculate a sum instead of an average, I assign a weight of one to every Queen neighbour and use `style = "B"`.

```{r}
#| label: binary-queen-weights
b_weights <- lapply(wm_q, function(x) 0 * x + 1)
b_weights2 <- nb2listw(wm_q, glist = b_weights, style = "B")
b_weights2
```

```{r}
#| label: lag-sum
lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
lag_sum
hunan <- left_join(hunan, lag.res, by = "NAME_3")
```

For `r hunan$County[1]`, the neighbour sum is **`r format(hunan[["lag_sum GDPPC"]][1], big.mark = ",", trim = TRUE)`**, equal to its neighbour average multiplied by `r card(wm_q)[1]`. This sum depends on both the neighbours' GDPPC values and the number of neighbours. Since GDPPC is a per-capita measure, adding it across counties does not give their combined GDP.

```{r}
#| label: lag-sum-maps
#| fig-cap: "County GDPPC and the sum of GDPPC values in neighbouring counties."
gdppc <- qtm(hunan, fill = "GDPPC")
lag_sum_gdppc <- qtm(hunan, fill = "lag_sum GDPPC")
tmap_arrange(gdppc, lag_sum_gdppc, asp = 1, ncol = 2)
```

### Spatial window average

The spatial window average includes the county itself as well as its neighbours. I add the diagonal with `include.self()` before row-standardising the weights.

```{r}
#| label: self-inclusive-weights
wm_qs <- include.self(wm_q)
wm_qs[[1]]
wm_qs <- nb2listw(wm_qs, style = "W")
wm_qs
```

The first county's window now contains `r length(wm_qs$neighbours[[1]])` counties, including itself.

```{r}
#| label: window-average
lag_w_avg_gpdpc <- lag.listw(wm_qs, hunan$GDPPC)
lag_w_avg_gpdpc

lag.list.wm_qs <- list(hunan$NAME_3, lag_w_avg_gpdpc)
lag_wm_qs.res <- as.data.frame(lag.list.wm_qs)
colnames(lag_wm_qs.res) <- c("NAME_3", "lag_window_avg GDPPC")
hunan <- left_join(hunan, lag_wm_qs.res, by = "NAME_3")
```

```{r}
#| label: window-average-table
hunan %>%
  st_drop_geometry() %>%
  select(County, `lag GDPPC`, `lag_window_avg GDPPC`) %>%
  kable(digits = 2)
```

Including the county's own GDPPC moves the window average towards that value. For `r hunan$County[1]`, the neighbour average is `r format(round(hunan[["lag GDPPC"]][1], 2), big.mark = ",", scientific = FALSE, trim = TRUE)`, while the window average is `r format(round(hunan[["lag_window_avg GDPPC"]][1], 2), big.mark = ",", scientific = FALSE, trim = TRUE)`.

For this comparison, I use the same class boundaries and colours on both maps so that the same colour represents the same range of values.

```{r}
#| label: window-average-maps
#| fig-cap: "Neighbour average and self-inclusive window average, using a common legend scale."
average_breaks <- pretty(range(c(hunan$`lag GDPPC`,
                                 hunan$`lag_window_avg GDPPC`)), n = 5)
lag_gdppc <- tm_shape(hunan) +
  tm_polygons(fill = "lag GDPPC",
              fill.scale = tm_scale_intervals(breaks = average_breaks,
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "Neighbour average"))
w_avg_gdppc <- tm_shape(hunan) +
  tm_polygons(fill = "lag_window_avg GDPPC",
              fill.scale = tm_scale_intervals(breaks = average_breaks,
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "Window average"))
tmap_arrange(lag_gdppc, w_avg_gdppc, asp = 1, ncol = 2)
```

### Spatial window sum

The window sum also includes the county itself, but uses weights of one without row standardisation.

```{r}
#| label: self-inclusive-binary-weights
wm_qs <- include.self(wm_q)
wm_qs
b_weights <- lapply(wm_qs, function(x) 0 * x + 1)
b_weights[1]
b_weights2 <- nb2listw(wm_qs, glist = b_weights, style = "B")
b_weights2
```

```{r}
#| label: window-sum
w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
hunan <- left_join(hunan, w_sum_gdppc.res, by = "NAME_3")
```

```{r}
#| label: window-sum-table
hunan %>%
  st_drop_geometry() %>%
  select(County, `lag_sum GDPPC`, `w_sum GDPPC`) %>%
  kable(digits = 0)
```

The difference between the window sum and the neighbour sum is exactly the county's own GDPPC. For `r hunan$County[1]`, adding its GDPPC of `r format(hunan$GDPPC[1], big.mark = ",", scientific = FALSE, trim = TRUE)` to the neighbour sum gives a window sum of `r format(hunan[["w_sum GDPPC"]][1], big.mark = ",", scientific = FALSE, trim = TRUE)`.

```{r}
#| label: window-sum-maps
#| fig-cap: "Neighbour sum and self-inclusive window sum, using a common legend scale."
sum_breaks <- pretty(range(c(hunan$`lag_sum GDPPC`,
                             hunan$`w_sum GDPPC`)), n = 5)
lag_sum_gdppc <- tm_shape(hunan) +
  tm_polygons(fill = "lag_sum GDPPC",
              fill.scale = tm_scale_intervals(breaks = sum_breaks,
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "Neighbour sum"))
w_sum_gdppc <- tm_shape(hunan) +
  tm_polygons(fill = "w_sum GDPPC",
              fill.scale = tm_scale_intervals(breaks = sum_breaks,
                                              values = "brewer.blues"),
              fill.legend = tm_legend(title = "Window sum"))
tmap_arrange(lag_sum_gdppc, w_sum_gdppc, asp = 1, ncol = 2)
```

```{r}
#| label: verify-spatial-calculations
#| include: false
neighbour_means <- vapply(wm_q, function(i) mean(hunan$GDPPC[i]), numeric(1))
neighbour_sums <- vapply(wm_q, function(i) sum(hunan$GDPPC[i]), numeric(1))
window_means <- (neighbour_sums + hunan$GDPPC) / (card(wm_q) + 1)
stopifnot(
  nrow(hunan) == 88L,
  !anyDuplicated(hunan$NAME_3),
  sum(card(wm_q)) == 448L,
  sum(card(wm_r)) == 440L,
  sum(card(wm_d62)) == 324L,
  all(card(wm_d62) > 0L),
  n_comp$nc == 1L,
  all(card(knn6) == 6L),
  all(abs(vapply(rswm_q$weights, sum, numeric(1)) - 1) < 1e-12),
  isTRUE(all.equal(hunan[["lag GDPPC"]], neighbour_means)),
  isTRUE(all.equal(hunan[["lag_sum GDPPC"]], neighbour_sums)),
  isTRUE(all.equal(hunan[["lag_window_avg GDPPC"]], window_means)),
  isTRUE(all.equal(hunan[["w_sum GDPPC"]], neighbour_sums + hunan$GDPPC))
)
```

## References

- Kam, T. S. [Spatial Weights and Applications](https://r4gdsa.netlify.app/chap08.html), *R for Geospatial Data Science and Analytics*. The analysis follows the supplied Chapter 8 teaching material.
- **spdep** documentation: [Creating neighbours using sf objects](https://r-spatial.github.io/spdep/articles/nb_sf.html).
