---
title: "Hands-on Exercise 4A: Funnel Plots for Fair Comparisons"
author: "Bao Sihan"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
execute:
  warning: false
  message: false
---

# 8.1 Overview

In this hands-on exercise, I practise **funnel plots** for fair comparison across sub-districts.\
I follow the structure in the teaching notes and create:

-   funnel plots using **FunnelPlotR**,
-   a static funnel plot built step-by-step with **ggplot2**, and
-   an interactive version using **plotly**.

# 8.2 Installing and launching R packages

```{r}
pacman::p_load(tidyverse, FunnelPlotR, plotly, knitr)
```

# 8.3 Importing data

```{r}
covid19 <- read_csv("../data/COVID-19_DKI_Jakarta.csv") %>%
  mutate_if(is.character, as.factor)

knitr::kable(head(covid19), format = "html")
```

# 8.4 FunnelPlotR methods

## 8.4.1 The basic plot

```{r}
funnel_plot(
  .data = covid19,
  numerator = Positive,
  denominator = Death,
  group = `Sub-district`
)
```

## 8.4.2 Makeover 1

Use `data_type = "PR"` to show proportions (Death / Positive), and zoom by `xrange` and `yrange`.

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",
  xrange = c(0, 6500),
  yrange = c(0, 0.05)
)
```

## 8.4.3 Makeover 2

Remove outlier labels and add titles.

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",
  xrange = c(0, 6500),
  yrange = c(0, 0.05),
  label = NA,
  title = "Cumulative COVID-19 Fatality Rate by Cumulative Total Number of COVID-19 Positive Cases",
  x_label = "Cumulative COVID-19 Positive Cases",
  y_label = "Cumulative Fatality Rate"
)
```

# 8.5 Funnel plot with ggplot2

## 8.5.1 Computing the derived fields

```{r}
df <- covid19 %>%
  mutate(rate = Death / Positive) %>%
  mutate(rate.se = sqrt((rate * (1 - rate)) / Positive)) %>%
  filter(rate > 0)

fit.mean <- weighted.mean(df$rate, 1 / df$rate.se^2)
```

## 8.5.2 Compute 95% and 99.9% control limits

```{r}
number.seq <- seq(1, max(df$Positive), 1)

number.ll95  <- fit.mean - 1.96 * sqrt((fit.mean * (1 - fit.mean)) / number.seq)
number.ul95  <- fit.mean + 1.96 * sqrt((fit.mean * (1 - fit.mean)) / number.seq)
number.ll999 <- fit.mean - 3.29 * sqrt((fit.mean * (1 - fit.mean)) / number.seq)
number.ul999 <- fit.mean + 3.29 * sqrt((fit.mean * (1 - fit.mean)) / number.seq)

dfCI <- data.frame(
  number.ll95, number.ul95, number.ll999, number.ul999,
  number.seq, fit.mean
)
```

## 8.5.3 Static funnel plot

```{r}
p <- ggplot(df, aes(x = Positive, y = rate)) +
  geom_point(aes(label = `Sub-district`), alpha = 0.4) +
  geom_line(data = dfCI, aes(x = number.seq, y = number.ll95),
            size = 0.4, colour = "grey40", linetype = "dashed") +
  geom_line(data = dfCI, aes(x = number.seq, y = number.ul95),
            size = 0.4, colour = "grey40", linetype = "dashed") +
  geom_line(data = dfCI, aes(x = number.seq, y = number.ll999),
            size = 0.4, colour = "grey40") +
  geom_line(data = dfCI, aes(x = number.seq, y = number.ul999),
            size = 0.4, colour = "grey40") +
  geom_hline(data = dfCI, aes(yintercept = fit.mean),
             size = 0.4, colour = "grey40") +
  coord_cartesian(ylim = c(0, 0.05)) +
  annotate("text", x = 1,   y = -0.13, label = "95%", size = 3, colour = "grey40") +
  annotate("text", x = 4.5, y = -0.18, label = "99%", size = 3, colour = "grey40") +
  ggtitle("Cumulative Fatality Rate by Cumulative Number of COVID-19 Cases") +
  xlab("Cumulative Number of COVID-19 Cases") +
  ylab("Cumulative Fatality Rate") +
  theme_light() +
  theme(
    plot.title = element_text(size = 12),
    legend.position = c(0.91, 0.85),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 7),
    legend.background = element_rect(colour = "grey60", linetype = "dotted"),
    legend.key.height = unit(0.3, "cm")
  )

p
```

## 8.5.4 Interactive funnel plot (plotly + ggplot2)

```{r}
fp_ggplotly <- ggplotly(
  p,
  tooltip = c("label", "x", "y")
)

fp_ggplotly
```
