---
title: "Hands-on Exercise 4C: Visual Statistical Analysis"
author: "Bao Sihan"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
execute:
  warning: false
  message: false
---

# 10.1 Overview

In this hands-on exercise, I practise **visual statistical analysis** using `ggstatsplot`. The plots include statistical test results directly in the visual.

# 10.2 Installing and launching R packages

```{r}
pacman::p_load(ggstatsplot, tidyverse)
```

# 10.3 Importing data

```{r}
exam <- read_csv("../data/Exam_data.csv")
glimpse(exam)
```

# 10.4 One-sample test: gghistostats()

```{r}
set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```

# 10.5 Two-sample mean test: ggbetweenstats()

```{r}
ggbetweenstats(
  data = exam,
  x = GENDER,
  y = MATHS,
  type = "np",
  messages = FALSE
)
```

# 10.6 One-way ANOVA: ggbetweenstats()

```{r}
ggbetweenstats(
  data = exam,
  x = RACE,
  y = ENGLISH,
  type = "p",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

# 10.7 Significant test of correlation: ggscatterstats()

```{r}
ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE
)
```

# 10.8 Significant test of association: ggbarstats()

```{r}
exam1 <- exam %>%
  mutate(
    MATHS_bins = cut(MATHS, breaks = c(0, 60, 75, 85, 100))
  )

ggbarstats(
  data = exam1,
  x = MATHS_bins,
  y = GENDER
)
```
