---
title: "In Class Exercise 6: Emerging Hot Spot Analysis"
author: "Bao Sihan"
date: "October 4, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show R code"
execute:
  echo: true
  results: hide
  warning: false
  message: false
  fig-width: 10
  fig-height: 6.5
---

## Overview

In this exercise, I investigate how clusters of GDP per capita (GDPPC) change across Hunan's counties from 2005 to 2021. Following [Chapter 11](https://r4gdsa.netlify.app/chap11.html#mann-kendall-test), I build a space-time cube, calculate yearly local Gi* statistics, test their trends using Mann–Kendall, and perform emerging hot spot analysis (EHSA).

Gi* describes the concentration of high or low values around a location. Mann–Kendall examines the direction of a series over time. EHSA combines the history of significant hot and cold spots with the trend in the statistic. A county's GDPPC level, its spatial clustering and its trend describe different aspects of the data.

The page presents notes and figures. The code is available in the expandable blocks.

## Getting Started

### Packages and data

I use the instructor's [Hunan GDPPC table](https://github.com/tskam/ISSS626-AY2026-27Aug/blob/master/In-class_Ex/In-class_Ex05/data/aspatial/Hunan_GDPPC.csv) and the Hunan county boundaries used in the earlier exercises. File paths are relative to this exercise folder.

```{r}
#| label: setup
pacman::p_load(sf, sfdep, tmap, plotly, tidyverse, Kendall)
tmap_mode("plot")
theme_set(theme_minimal(base_size = 11))
spdep::set.coresOption(1L)
```

### Importing and checking the panel

I transform the boundaries to UTM zone 50N so that distances are measured in metres. County names provide the link between the geometry and the annual records. I explicitly sort both datasets before creating the cube, preserving the correspondence between counties, weights and results.

```{r}
#| label: import-data
hunan <- st_read("data/geospatial/Hunan.shp", quiet = TRUE) %>%
  st_transform(32650) %>%
  arrange(County)
GDPPC <- read_csv("data/aspatial/Hunan_GDPPC.csv", show_col_types = FALSE) %>%
  mutate(Year = as.integer(Year)) %>%
  arrange(Year, County)
years <- sort(unique(GDPPC$Year))
```

```{r}
#| label: validate-panel
#| include: false
stopifnot(nrow(hunan) == 88L, !anyDuplicated(hunan$County),
          all(st_is_valid(hunan)), st_crs(hunan)$epsg == 32650,
          nrow(GDPPC) == 1496L, length(years) == 17L,
          identical(years, 2005:2021),
          !anyNA(GDPPC), all(is.finite(GDPPC$GDPPC)),
          all(GDPPC$GDPPC > 0),
          !anyDuplicated(GDPPC[c("County", "Year")]),
          setequal(GDPPC$County, hunan$County),
          all(table(GDPPC$County) == length(years)),
          all(table(GDPPC$Year) == nrow(hunan)))
```

The checks confirm **`r nrow(hunan)` counties**, **`r length(years)` annual time steps** and **`r nrow(GDPPC)` county-year observations**, with no missing values or duplicate county-year keys. Every county has an observation in every year.

```{r}
#| label: panel-distribution
#| fig-cap: "GDPPC distributions across the 88 counties in each year."
ggplot(GDPPC, aes(factor(Year), GDPPC)) +
  geom_boxplot(fill = "#b9d8ea", outlier.alpha = 0.35) +
  labs(x = "Year", y = "GDP per capita (source units)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The distributions show the county-level spread within each year. These are the supplied GDPPC values; the exercise does not apply an inflation adjustment.

## Creating a Space-Time Cube

The cube stores one boundary for each county and one GDPPC record for each county-year combination. A complete cube is essential because the later calculations use the same set of locations at each time step.

```{r}
#| label: create-cube
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County", .time_col = "Year")
stopifnot(is_spacetime_cube(GDPPC_st))
```

```{r}
#| label: endpoint-gdppc-maps
#| fig-cap: "The first and last years of the GDPPC panel, using a shared colour scale."
endpoint_data <- hunan %>%
  left_join(filter(GDPPC, Year %in% range(years)),
            by = "County", relationship = "one-to-many")
ggplot(endpoint_data) +
  geom_sf(aes(fill = GDPPC), colour = "white", linewidth = 0.1) +
  facet_wrap(~ Year, nrow = 1) +
  scale_fill_viridis_c(option = "C", name = "GDPPC") +
  coord_sf(datum = NA) + theme_void()
```

A common colour scale makes the two years comparable. These maps show the original values; the hot spot analysis evaluates their spatial concentration.

## Computing Yearly Gi*

### Neighbours and spatial weights

I use Queen contiguity: counties that share an edge or a vertex are neighbours. Following the chapter, I include each county in its neighbour list and calculate inverse-distance weights between representative points inside the polygons. Nearby neighbours receive greater weight.

```{r}
#| label: yearly-weights
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(st_contiguity(geometry)),
         wt = st_inverse_distance(nb, geometry, scale = 1, alpha = 1),
         .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")
queen_nb <- st_contiguity(st_geometry(hunan))
stopifnot(all(lengths(queen_nb) > 0))
```

In sfdep, the zero-distance self link receives weight zero under inverse-distance weighting. I retain this chapter setting for the yearly demonstration. The final EHSA calculation uses its default row-standardised Queen weights, including positive self weights, and adds temporal neighbours.

```{r}
#| label: neighbour-map
#| fig-cap: "Queen neighbours of the Hunan counties. Links join representative points within the polygons."
centres <- st_coordinates(st_point_on_surface(st_geometry(hunan)))
edges <- tibble(from = rep(seq_along(queen_nb), lengths(queen_nb)),
                to = unlist(queen_nb)) %>% filter(from < to)
ggplot() +
  geom_sf(data = hunan, fill = "#f3f5f7", colour = "#9aa3ac", linewidth = 0.2) +
  geom_segment(data = edges,
               aes(x = centres[from, 1], y = centres[from, 2],
                   xend = centres[to, 1], yend = centres[to, 2]),
               colour = "#54728a", linewidth = 0.35, alpha = 0.7) +
  geom_point(aes(centres[, 1], centres[, 2]), size = 0.6, colour = "#23435b") +
  coord_sf(datum = NA) + theme_void()
```

### Calculating the statistic for each year

I calculate the spatial statistic separately for each year, using 499 conditional permutations and a fixed seed. The county order within every year matches the boundary order.

```{r}
#| label: yearly-gstar
gi_stars <- GDPPC_nb %>%
  activate("data") %>%
  as_tibble() %>%
  group_by(Year) %>%
  mutate(gi_star = local_gstar_perm(GDPPC, nb, wt,
                                   nsim = 499, iseed = 62606)) %>%
  unnest(gi_star) %>%
  ungroup() %>%
  arrange(County, Year)
stopifnot(nrow(gi_stars) == nrow(GDPPC),
          all(is.finite(gi_stars$gi_star)))
```

```{r}
#| label: yearly-gstar-maps
#| fig-cap: "Yearly Gi* values at the beginning and end of the panel. These maps display the statistic without a significance mask."
gi_endpoints <- hunan %>%
  left_join(gi_stars %>% filter(Year %in% range(years)) %>%
              select(County, Year, gi_star),
            by = "County", relationship = "one-to-many")
gi_limit <- max(abs(gi_endpoints$gi_star))
ggplot(gi_endpoints) +
  geom_sf(aes(fill = gi_star), colour = "white", linewidth = 0.1) +
  facet_wrap(~ Year, nrow = 1) +
  scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#b2182b",
                       limits = c(-gi_limit, gi_limit), name = "Gi*") +
  coord_sf(datum = NA) + theme_void()
```

Positive values indicate concentration of relatively high GDPPC, while negative values indicate concentration of relatively low GDPPC. The colours alone do not establish statistical significance.

## Mann–Kendall Trend Test

### Changsha example

Mann–Kendall tests for a monotonic trend without requiring normally distributed observations. Its null hypothesis is that there is no monotonic trend. Kendall's tau ranges from −1 to 1: a positive value indicates an increasing tendency and a negative value indicates a decreasing tendency. Individual years can still fluctuate.

```{r}
#| label: changsha-trend
#| fig-cap: "Changsha's yearly Gi* series under the chapter's inverse-distance weights."
cbg <- gi_stars %>%
  filter(County == "Changsha") %>%
  arrange(Year) %>% select(County, Year, gi_star)
changsha_mk <- Kendall::MannKendall(cbg$gi_star)
p <- ggplot(cbg, aes(Year, gi_star)) +
  geom_hline(yintercept = 0, colour = "grey65", linetype = "dashed") +
  geom_line(colour = "#245a81", linewidth = 0.8) +
  geom_point(colour = "#245a81", size = 2) +
  scale_x_continuous(breaks = years[seq(1, length(years), 2)]) +
  labs(x = "Year", y = "Gi*", title = "Changsha: spatial clustering over time")
p
```

For Changsha, **tau = `r sprintf('%.3f', changsha_mk$tau)`** and **p = `r format.pval(changsha_mk$sl, digits = 3, eps = 0.0001)`**. At the 5% level, this result `r if (changsha_mk$sl < 0.05) 'supports a statistically significant monotonic trend' else 'does not provide sufficient evidence of a monotonic trend'`. The direction is `r if (changsha_mk$tau > 0) 'upward' else if (changsha_mk$tau < 0) 'downward' else 'flat'` in the spatial statistic.

### Interactive version

Hover over a point to inspect its year and Gi* value. The interactive view uses the same series as the static figure.

```{r}
#| label: interactive-changsha
#| results: asis
ggplotly(p, tooltip = c("x", "y")) %>%
  layout(autosize = TRUE, margin = list(l = 55, r = 15, t = 55, b = 45)) %>%
  config(displaylogo = FALSE)
```

### Trends across all counties

I apply the same test to each county's chronologically ordered yearly Gi* series. The reported p-values are unadjusted. A significant upward trend can describe an intensifying positive statistic or a negative statistic moving towards zero, so its sign must be read together with the series.

```{r}
#| label: county-trends
mk_results <- gi_stars %>%
  arrange(County, Year) %>%
  group_by(County) %>%
  summarise(mk = list(unclass(Kendall::MannKendall(gi_star))),
            .groups = "drop") %>%
  unnest_wider(mk) %>%
  mutate(trend = case_when(
    sl >= 0.05 ~ "No significant trend",
    tau > 0 ~ "Increasing Gi*",
    TRUE ~ "Decreasing Gi*"))
trend_levels <- c("Decreasing Gi*", "No significant trend", "Increasing Gi*")
trend_colours <- c("Decreasing Gi*" = "#2166ac",
                   "No significant trend" = "#d9d9d9",
                   "Increasing Gi*" = "#b2182b")
trend_sf <- hunan %>% left_join(mk_results, by = "County")
```

```{r}
#| label: county-trend-map
#| fig-cap: "Mann–Kendall trends in the yearly Gi* statistic, using an unadjusted 5% significance level."
ggplot(trend_sf) +
  geom_sf(aes(fill = factor(trend, levels = trend_levels)),
          colour = "white", linewidth = 0.15) +
  scale_fill_manual(values = trend_colours, name = NULL, drop = FALSE) +
  coord_sf(datum = NA) + theme_void() +
  theme(legend.position = "bottom")
```

There are **`r sum(mk_results$sl < 0.05 & mk_results$tau > 0)` increasing** and **`r sum(mk_results$sl < 0.05 & mk_results$tau < 0)` decreasing** significant Gi* trends. The remaining **`r sum(mk_results$sl >= 0.05)` counties** do not pass this trend threshold.

```{r}
#| label: strongest-trends
#| fig-cap: "The ten counties with the smallest Mann–Kendall p-values; ties are ordered by absolute tau."
strongest <- mk_results %>%
  arrange(sl, desc(abs(tau)), County) %>% slice_head(n = 10)
ggplot(strongest, aes(reorder(County, tau), tau, fill = trend)) +
  geom_col(width = 0.7) + coord_flip() +
  scale_fill_manual(values = trend_colours, guide = "none") +
  scale_y_continuous(limits = c(-1, 1)) +
  labs(x = NULL, y = "Kendall's tau")
```

## Emerging Hot Spot Analysis

### Space-time neighbours and classification

I follow the chapter's final call with a one-step temporal lag and 99 simulations. With no custom weight columns supplied, sfdep constructs row-standardised Queen weights including each county itself. I explicitly retain the function's 1% classification threshold and set the random seed for reproducibility.

```{r}
#| label: compute-ehsa
set.seed(62606)
ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st, .var = "GDPPC", k = 1, nsim = 99,
  threshold = 0.01, include_gi = TRUE)
ehsa_gi <- attr(ehsa, "gi_star")
ehsa$classification_raw <- ehsa$classification
ehsa$classification <- str_replace_all(ehsa$classification,
  c("oscilating" = "oscillating", "hotspot" = "hot spot", "coldspot" = "cold spot"))
hunan_ehsa <- hunan %>%
  left_join(as_tibble(ehsa), by = c("County" = "location"))
```

The classification uses the sequence of significant positive and negative space-time statistics and their trend. In this implementation, the returned `p_value` is the Mann–Kendall trend p-value. The function does not apply a false discovery rate correction automatically. With 99 simulations, the smallest simulated p-value is 0.01, so the classification is an exploratory result at relatively coarse permutation resolution.

```{r}
#| label: ehsa-class-distribution
#| fig-cap: "Distribution of the EHSA classes across all 88 counties."
class_counts <- as_tibble(ehsa) %>% count(classification, sort = TRUE)
ggplot(class_counts, aes(reorder(classification, n), n)) +
  geom_col(fill = "#486b83", width = 0.7) +
  geom_text(aes(label = n), hjust = -0.15) + coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.13))) +
  labs(x = NULL, y = "Counties")
```

The most frequent class is **`r class_counts$classification[1]`**, containing **`r class_counts$n[1]` counties**. Class labels describe the history of spatial clustering rather than GDPPC growth itself.

### Mapping all classifications

I show the classifications for all counties before applying the chapter's additional trend filter. This preserves classes whose definition does not require a significant monotonic trend.

```{r}
#| label: ehsa-map
#| fig-height: 8
#| fig-cap: "EHSA classifications for all Hunan counties."
all_class_colours <- c(
  "no pattern detected" = "#d9d9d9",
  "new hot spot" = "#fee08b", "consecutive hot spot" = "#fdae61",
  "intensifying hot spot" = "#a50026", "persistent hot spot" = "#d73027",
  "diminishing hot spot" = "#f46d43", "sporadic hot spot" = "#fdbf6f",
  "oscillating hot spot" = "#e78ac3", "historical hot spot" = "#bf812d",
  "new cold spot" = "#d1e5f0", "consecutive cold spot" = "#92c5de",
  "intensifying cold spot" = "#053061", "persistent cold spot" = "#2166ac",
  "diminishing cold spot" = "#4393c3", "sporadic cold spot" = "#80b1d3",
  "oscillating cold spot" = "#8073ac", "historical cold spot" = "#35978f")
observed_classes <- sort(unique(ehsa$classification))
stopifnot(all(observed_classes %in% names(all_class_colours)))
class_colours <- all_class_colours[observed_classes]
ehsa_map <- tm_shape(hunan_ehsa) +
  tm_polygons(fill = "classification",
              fill.scale = tm_scale_categorical(values = class_colours),
              fill.legend = tm_legend(title = "EHSA class")) +
  tm_borders(col = "white", lwd = 0.4) +
  tm_title("Emerging hot spot analysis: 2005-2021")
ehsa_map
```

### The chapter's trend-significance view

The chapter also filters the output to counties with a Mann–Kendall p-value below 0.05. Grey counties here are excluded by that trend filter. Their colour in the previous map remains the full EHSA classification.

```{r}
#| label: ehsa-trend-mask
#| fig-height: 8
#| fig-cap: "EHSA classifications retained by the chapter's p_value < 0.05 trend filter."
ehsa_sig <- hunan_ehsa %>% filter(p_value < 0.05)
tm_shape(hunan_ehsa) +
  tm_polygons(fill = "#e6e6e6", col = "white", lwd = 0.4) +
  tm_shape(ehsa_sig) +
  tm_polygons(fill = "classification",
              fill.scale = tm_scale_categorical(values = class_colours),
              fill.legend = tm_legend(title = "EHSA class")) +
  tm_borders(col = "white", lwd = 0.4) +
  tm_title("EHSA classes with a significant Gi* trend")
```

### Reading the class histories

A **new** spot occurs only at the final time step. A **consecutive** spot has an uninterrupted run ending at the final step, with fewer than 90% of all steps significant. **Persistent**, **intensifying** and **diminishing** classes have significant spots in at least 90% of the steps; their labels distinguish the trend in the strength of clustering.

**Sporadic** spots have intermittent significance. **Oscillating** spots have experienced significance of both signs, while **historical** spots have a long history of significance but do not qualify as a spot at the final step. **No pattern detected** means that the sequence does not satisfy an EHSA class rule.

The following examples plot one county from each observed class, up to six classes. The points highlight the significant space-time statistics used by EHSA, rather than the standalone yearly statistics above.

```{r}
#| label: ehsa-class-examples
#| fig-height: 8
#| fig-cap: "Example histories from the observed EHSA classes. Filled points meet the 1% simulation threshold."
ehsa_series <- GDPPC %>%
  mutate(gi_star = ehsa_gi$gi_star, p_sim = ehsa_gi$p_sim) %>%
  left_join(as_tibble(ehsa) %>% select(location, classification, tau),
            by = c("County" = "location"))
examples <- as_tibble(ehsa) %>%
  arrange(classification, desc(abs(tau)), location) %>%
  group_by(classification) %>% slice_head(n = 1) %>% ungroup() %>%
  slice_head(n = 6)
ehsa_series %>% filter(County %in% examples$location) %>%
  mutate(panel = paste(County, classification, sep = "\n")) %>%
  ggplot(aes(Year, gi_star)) +
  geom_hline(yintercept = 0, colour = "grey65", linetype = "dashed") +
  geom_line(colour = "#486b83") +
  geom_point(aes(shape = p_sim <= 0.01), size = 2, colour = "#245a81") +
  scale_shape_manual(values = c(`FALSE` = 1, `TRUE` = 16),
                     name = "Significant bin") +
  facet_wrap(~ panel, ncol = 2, scales = "free_y") +
  labs(x = "Year", y = "Space-time Gi*") +
  theme(legend.position = "bottom")
```

## Interpretation

The annual Gi* series reveals how each county's spatial concentration changes relative to other counties in the same year. Mann–Kendall summarises its monotonic direction. EHSA adds the timing and frequency of significant hot and cold spots, using its own space-time neighbourhood.

This exercise uses a fixed neighbourhood and the supplied GDPPC series. The unadjusted significance levels and 99-simulation EHSA setting follow the teaching workflow. The ordinary Mann–Kendall test assumes independent observations in time; serial dependence can affect its p-values. A substantive economic study would also examine price comparability, temporal dependence, sensitivity to neighbourhood choices and a larger number of simulations.

```{r}
#| label: validate-and-save
#| include: false
stopifnot(nrow(mk_results) == nrow(hunan),
          nrow(ehsa) == nrow(hunan),
          identical(as.character(ehsa$location), hunan$County),
          !anyNA(ehsa$classification),
          !anyNA(hunan_ehsa$tau),
          all(is.finite(ehsa$p_value)),
          all(ehsa$p_value >= 0 & ehsa$p_value <= 1),
          length(ehsa_gi$gi_star) == nrow(GDPPC),
          all(is.finite(ehsa_gi$gi_star)),
          sum(class_counts$n) == nrow(hunan))
for (yr in years) {
  stopifnot(identical(GDPPC$County[GDPPC$Year == yr], hunan$County))
}
dir.create("data/rds", showWarnings = FALSE)
write_rds(list(panel = GDPPC, geometry = hunan,
               yearly_gi = gi_stars, mann_kendall = mk_results,
               ehsa = ehsa, ehsa_series = ehsa_series,
               settings = list(seed = 62606, yearly_nsim = 499,
                               ehsa_nsim = 99, k = 1, threshold = 0.01,
                               sfdep_version = as.character(packageVersion("sfdep")))),
          "data/rds/analysis_results.rds")
```

## References

[Chapter 11: Emerging Hot Spot Analysis](https://r4gdsa.netlify.app/chap11.html), [sfdep source and documentation](https://github.com/JosiahParry/sfdep), [Mann–Kendall test guidance](https://vsp.pnnl.gov/help/vsample/design_trend_mann_kendall.htm) and the [instructor's data repository](https://github.com/tskam/ISSS626-AY2026-27Aug).
