---
title: "Hands-on Exercise 10: Information Dashboard Design"
author: "Bao Sihan"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This hands-on exercise is based on Chapter 31, which focuses on information dashboard design in R. In this page, I work on three main ideas: bullet charts, sparklines, and dashboard-style tables.

To make the file easier to render in my local environment, I use the prepared `CoffeeChain.rds` file saved in `../data/` instead of importing the original Access database.

The goals of this exercise are:

-   to build a bullet chart for product sales comparison,
-   to create sparkline views for monthly sales trends,
-   to combine charts and tables into compact dashboard-style outputs,
-   to create an interactive table with hover values in a stable way.

## Getting started

### Packages used

```{r}
required_pkgs <- c(
  "tidyverse", "lubridate", "ggthemes", "gt",
  "reactable", "htmltools", "scales"
)

optional_pkgs <- c("gtExtras")

missing_required <- required_pkgs[
  !vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)
]

missing_optional <- optional_pkgs[
  !vapply(optional_pkgs, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_required) > 0) {
  install.packages(missing_required, repos = "https://cloud.r-project.org")
}

if (length(missing_optional) > 0) {
  try(
    install.packages(missing_optional, repos = "https://cloud.r-project.org"),
    silent = TRUE
  )
}

invisible(lapply(required_pkgs, library, character.only = TRUE))

has_gtextras <- requireNamespace("gtExtras", quietly = TRUE)
if (has_gtextras) {
  library(gtExtras)
}
```

### Importing the prepared data

```{r}
coffeechain <- readr::read_rds("../data/CoffeeChain.rds") %>%
  mutate(Date = as.Date(Date))
```

### Preparing product-level totals

```{r}
product <- coffeechain %>%
  group_by(Product) %>%
  summarise(
    target = sum(`Budget Sales`, na.rm = TRUE),
    current = sum(Sales, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Product = forcats::fct_reorder(Product, current))
```

### Preparing the 2013 monthly report

```{r}
report_base <- coffeechain %>%
  mutate(Year = year(Date), Month = month(Date)) %>%
  filter(Year == 2013) %>%
  group_by(Product, Month) %>%
  summarise(
    Sales = sum(Sales, na.rm = TRUE),
    `Budget Sales` = sum(`Budget Sales`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  tidyr::complete(
    Product,
    Month = 1:12,
    fill = list(Sales = 0, `Budget Sales` = 0)
  ) %>%
  arrange(Product, Month)
```

## Bullet chart in ggplot2

The chart below recreates the product-level bullet chart idea from the chapter. The grey bars provide reference ranges, the black bar shows actual sales, and the red marker shows the target value.

```{r}
bullet_data <- product %>%
  mutate(
    range3 = pmax(target, current) * 1.05,
    range2 = target * 0.80,
    range1 = target * 0.50
  )

ggplot(bullet_data, aes(x = Product)) +
  geom_col(aes(y = range3), fill = "grey85", width = 0.82) +
  geom_col(aes(y = range2), fill = "grey70", width = 0.82) +
  geom_col(aes(y = range1), fill = "grey55", width = 0.82) +
  geom_col(aes(y = current), fill = "black", width = 0.32) +
  geom_errorbar(
    aes(y = target, ymin = target, ymax = target),
    colour = "red",
    width = 0.38,
    linewidth = 1
  ) +
  coord_flip() +
  labs(
    title = "Bullet Chart of Product Sales",
    x = NULL,
    y = "Sales"
  ) +
  theme_minimal()
```

Observation: The black bars and red target markers make the performance comparison very direct.

Insight: A bullet chart is a good dashboard choice because it is compact and allows many products to be compared in one panel.

## Plotting sparklines using ggplot2

### Preparing sparkline support data

```{r}
sales_report <- report_base %>%
  select(Month, Product, Sales)

mins <- sales_report %>%
  group_by(Product) %>%
  slice_min(order_by = Sales, n = 1, with_ties = FALSE) %>%
  ungroup()

maxs <- sales_report %>%
  group_by(Product) %>%
  slice_max(order_by = Sales, n = 1, with_ties = FALSE) %>%
  ungroup()

ends <- sales_report %>%
  group_by(Product) %>%
  filter(Month == max(Month)) %>%
  ungroup()

quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(
    quart1 = quantile(Sales, 0.25, na.rm = TRUE),
    quart2 = quantile(Sales, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  right_join(sales_report, by = "Product")
```

### Building the ggplot sparkline panel

```{r}
ggplot(sales_report, aes(x = Month, y = Sales)) +
  facet_grid(Product ~ ., scales = "free_y") +
  geom_ribbon(
    data = quarts,
    aes(ymin = quart1, ymax = quart2),
    fill = "grey92"
  ) +
  geom_line(linewidth = 0.35, colour = "grey35") +
  geom_point(data = mins, colour = "red", size = 0.9) +
  geom_point(data = maxs, colour = "blue", size = 0.9) +
  geom_text(data = mins, aes(label = Sales), vjust = -1, size = 2) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5, size = 2) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.45, size = 2) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0, size = 2.2) +
  expand_limits(x = max(sales_report$Month) + 3) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(expand = c(0.1, 0)) +
  ggthemes::theme_tufte(base_size = 7, base_family = "sans") +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    strip.text = element_blank()
  )
```

Observation: The sparkline panel shows each product's monthly movement in a very small space.

Insight: This design is useful when the dashboard needs to emphasise trend instead of only one final total.

## Static information dashboard design: gt and gtExtras methods

### Creating the list-column for monthly sales

```{r}
report <- report_base %>%
  group_by(Product) %>%
  summarise(
    `Monthly Sales` = list(Sales),
    .groups = "drop"
  )
```

### A simple bullet chart table

```{r}
if (has_gtextras) {
  product %>%
    gt() %>%
    gt_plt_bullet(
      column = current,
      target = target,
      width = 60,
      palette = c("lightblue", "black")
    ) %>%
    gt_theme_538()
} else {
  product %>%
    gt() %>%
    fmt_number(columns = c(current, target), decimals = 0)
}
```

Observation: Embedding the bullet chart inside a table makes the output look more like a real dashboard.

Insight: This format is useful for management reporting because the chart and labels stay together in one place.

### Basic sparkline table

```{r}
if (has_gtextras) {
  report %>%
    gt() %>%
    gt_plt_sparkline(`Monthly Sales`, same_limit = FALSE)
} else {
  report %>%
    mutate(`Monthly Sales` = purrr::map_chr(`Monthly Sales`, ~ paste(.x, collapse = ", "))) %>%
    gt()
}
```

### Adding summary statistics

```{r}
sales <- report_base %>%
  group_by(Product) %>%
  summarise(
    Min = min(Sales, na.rm = TRUE),
    Max = max(Sales, na.rm = TRUE),
    Average = mean(Sales, na.rm = TRUE),
    .groups = "drop"
  )

sales %>%
  gt() %>%
  fmt_number(columns = Average, decimals = 2)
```

### Combining statistics and sparklines

```{r}
sales_data <- left_join(sales, report, by = "Product")

if (has_gtextras) {
  sales_data %>%
    gt() %>%
    gt_plt_sparkline(`Monthly Sales`, same_limit = FALSE) %>%
    fmt_number(columns = Average, decimals = 2)
} else {
  sales_data %>%
    mutate(`Monthly Sales` = purrr::map_chr(`Monthly Sales`, ~ paste(.x, collapse = ", "))) %>%
    gt() %>%
    fmt_number(columns = Average, decimals = 2)
}
```

### Combining sparklines and bullet charts

```{r}
bullet <- report_base %>%
  group_by(Product) %>%
  summarise(
    Target = sum(`Budget Sales`, na.rm = TRUE),
    Actual = sum(Sales, na.rm = TRUE),
    .groups = "drop"
  )

sales_data2 <- left_join(sales_data, bullet, by = "Product")

if (has_gtextras) {
  sales_data2 %>%
    gt() %>%
    gt_plt_sparkline(`Monthly Sales`) %>%
    gt_plt_bullet(
      column = Actual,
      target = Target,
      width = 28,
      palette = c("lightblue", "black")
    ) %>%
    fmt_number(columns = Average, decimals = 2) %>%
    gt_theme_538()
} else {
  sales_data2 %>%
    mutate(`Monthly Sales` = purrr::map_chr(`Monthly Sales`, ~ paste(.x, collapse = ", "))) %>%
    gt() %>%
    fmt_number(columns = Average, decimals = 2)
}
```

Observation: This combined table is much closer to a practical dashboard because it brings trend, level, and target information together.

Insight: A compact dashboard often works best when multiple small views are placed in one aligned table.

## Interactive information dashboard design: reactable methods

The original chapter uses `reactablefmtr`, but that package is not always easy to install on every R version. To keep this file stable and still preserve hover values, I create the interactive section with `reactable` and custom HTML mini charts.

### Helper functions for custom hover charts

```{r}
make_sparkbar <- function(values, labels = month.abb[seq_along(values)]) {
  values <- as.numeric(values)
  vmax <- max(values, na.rm = TRUE)
  vmean <- mean(values, na.rm = TRUE)
  min_id <- which.min(values)[1]
  max_id <- which.max(values)[1]

  mean_y <- if (vmax == 0) 0 else round(vmean / vmax * 26, 1)

  bars <- lapply(seq_along(values), function(i) {
    fill_col <- if (i == min_id) {
      "#d62728"
    } else if (i == max_id) {
      "#1f77b4"
    } else {
      "#9aa5b1"
    }

    bar_height <- if (vmax == 0) 2 else max(2, round(values[i] / vmax * 26))

    htmltools::tags$div(
      title = paste0(labels[i], ": ", scales::comma(values[i])),
      style = paste0(
        "flex:1; height:", bar_height,
        "px; background:", fill_col,
        "; border-radius:1px;"
      )
    )
  })

  htmltools::tags$div(
    style = "position:relative; width:230px; height:34px; padding-top:4px;",
    htmltools::tags$div(
      style = paste0(
        "position:absolute; left:0; right:0; bottom:",
        3 + mean_y,
        "px; border-top:1px dashed #d62728; opacity:0.9;"
      )
    ),
    htmltools::tags$div(
      style = "position:absolute; left:0; right:0; bottom:3px; display:flex; align-items:flex-end; gap:2px; height:26px;",
      htmltools::tagList(bars)
    ),
    htmltools::tags$div(
      style = "position:absolute; left:0; right:0; bottom:0; border-top:1px solid #d8d8d8;"
    )
  )
}

make_sparkline <- function(values, labels = month.abb[seq_along(values)]) {
  values <- as.numeric(values)
  n <- length(values)
  width <- 230
  height <- 42
  pad <- 7

  minv <- min(values, na.rm = TRUE)
  maxv <- max(values, na.rm = TRUE)
  rng <- maxv - minv
  if (rng == 0) rng <- 1

  xs <- seq(pad, width - pad, length.out = n)
  ys <- height - pad - ((values - minv) / rng) * (height - 2 * pad)
  mean_y <- height - pad - ((mean(values, na.rm = TRUE) - minv) / rng) * (height - 2 * pad)

  point_string <- paste(sprintf("%.1f,%.1f", xs, ys), collapse = " ")

  circle_string <- paste(
    vapply(seq_len(n), function(i) {
      fill_col <- if (i == which.min(values)[1]) {
        "#d62728"
      } else if (i == which.max(values)[1] || i == n) {
        "#1f77b4"
      } else {
        "#000000"
      }

      opacity_val <- if (i %in% c(which.min(values)[1], which.max(values)[1], n)) {
        "0.95"
      } else {
        "0.001"
      }

      paste0(
        '<circle cx="', round(xs[i], 1),
        '" cy="', round(ys[i], 1),
        '" r="3.2" fill="', fill_col,
        '" fill-opacity="', opacity_val,
        '"><title>', labels[i], ': ', scales::comma(values[i]), '</title></circle>'
      )
    }, character(1)),
    collapse = ""
  )

  svg_txt <- paste0(
    '<svg xmlns="http://www.w3.org/2000/svg" width="', width,
    '" height="', height,
    '" viewBox="0 0 ', width, ' ', height, '">',
    '<line x1="', pad, '" x2="', width - pad,
    '" y1="', round(mean_y, 1), '" y2="', round(mean_y, 1),
    '" stroke="#d62728" stroke-width="1" stroke-dasharray="3 2"/>',
    '<polyline fill="none" stroke="#6b7f99" stroke-width="1.3" points="', point_string, '"/>',
    circle_string,
    '<text x="0" y="13" font-size="10" fill="#444">', scales::comma(values[1]), '</text>',
    '<text x="', width - 30, '" y="13" font-size="10" fill="#444">', scales::comma(values[n]), '</text>',
    '</svg>'
  )

  htmltools::HTML(svg_txt)
}
```

### Preparing the interactive report table

```{r}
report_interactive <- report_base %>%
  group_by(Product) %>%
  summarise(
    `Monthly Sales` = list(Sales),
    Average = mean(Sales, na.rm = TRUE),
    Total = sum(Sales, na.rm = TRUE),
    .groups = "drop"
  )
```

### Interactive sparkbar table with hover values

```{r}
reactable(
  report_interactive,
  defaultPageSize = 13,
  compact = TRUE,
  striped = TRUE,
  highlight = TRUE,
  searchable = FALSE,
  columns = list(
    Product = colDef(minWidth = 170),
    `Monthly Sales` = colDef(
      html = TRUE,
      minWidth = 250,
      cell = function(value) make_sparkbar(unlist(value))
    ),
    Average = colDef(format = colFormat(separators = TRUE, digits = 1)),
    Total = colDef(format = colFormat(separators = TRUE, digits = 0))
  )
)
```

Observation: This view is close to a dashboard strip chart, and the monthly values can be seen by moving the mouse over each small bar.

Insight: This approach is more stable than depending on extra GitHub-only packages, so it is practical for assignment rendering.

### Interactive sparkline table with hover values

```{r}
reactable(
  report_interactive,
  defaultPageSize = 13,
  compact = TRUE,
  striped = TRUE,
  highlight = TRUE,
  searchable = FALSE,
  columns = list(
    Product = colDef(minWidth = 170),
    `Monthly Sales` = colDef(
      name = "Sales Trend",
      html = TRUE,
      minWidth = 250,
      cell = function(value) make_sparkline(unlist(value))
    ),
    Average = colDef(format = colFormat(separators = TRUE, digits = 1)),
    Total = colDef(format = colFormat(separators = TRUE, digits = 0))
  )
)
```

Observation: The sparkline version is better for reading the shape of the trend, while the red dashed line adds a useful reference for the mean level.

Insight: Compared with the sparkbar view, this design is more suitable when the user wants to focus on fluctuations over time.

## Conclusion

This exercise shows that dashboard design is not only about producing one attractive chart. A better dashboard usually combines compact comparisons, trend summaries, and table layout in the same page.

From this practice, I found that `ggplot2` is useful for building chart logic clearly, `gt` and `gtExtras` are useful for static dashboard tables, and `reactable` is helpful for creating lightweight interactive dashboard views. In my final version, I used a stable custom `reactable` solution so the file can render more smoothly while still keeping hover values.

## Reference

-   R for Visual Analytics, Chapter 31: Information Dashboard Design: R methods.
