---
title: "Hands-on Exercise 6: Geographical Segmentation with Spatially Constrained Clustering"
author: "Bao Sihan"
date: "October 3, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show R code"
execute:
  echo: true
  results: hide
  warning: false
  message: false
  fig-width: 11
  fig-height: 6.5
---

## Overview

In this exercise, I group the townships of Shan State, Myanmar, using household access to information and communication technology (ICT). I first explore the 2014 census indicators, then compare hierarchical clustering, SKATER and ClustGeo. The main question is whether townships with similar ICT profiles also form useful geographical regions.

The analysis follows [Chapter 12](https://r4gdsa.netlify.app/chap12.html). Each section presents the main idea and its figures; the R code is available in the expandable blocks.

## Getting Started

### Data and tools

The supplied data contain Myanmar township boundaries and six household ICT indicators for Shan State. I use the 55 townships in Shan (East), Shan (North) and Shan (South). The boundary layer and census table are joined using the township code, rather than the township name.

```{r}
#| label: load-packages
pacman::p_load(sf, spdep, tmap, ClustGeo, ggpubr, cluster,
               factoextra, heatmaply, corrplot, tidyverse, GGally)
tmap_mode("plot")
theme_set(theme_minimal(base_size = 11))
```

## Data Import and Preparation

### Importing the data

All paths are relative to this exercise folder. I keep the boundary order throughout the analysis so that each cluster label returns to the correct township.

```{r}
#| label: import-data
shan_sf <- st_read("data/geospatial/myanmar_township_boundaries.shp",
                   quiet = TRUE) %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(ST, ST_PCODE, DT, DT_PCODE, TS, TS_PCODE)

ict <- read_csv("data/aspatial/Shan-ICT.csv", show_col_types = FALSE)
```

### Deriving penetration rates

Raw ownership counts are affected by the number of households in a township. To compare access, I calculate the number of households owning each item per 1,000 households:

$$
\text{Penetration rate} =
\frac{\text{Households with the ICT item}}{\text{Total households}}\times 1{,}000.
$$

```{r}
#| label: derive-rates
ict_derived <- ict %>%
  rename(DT_PCODE = `District Pcode`, DT = `District Name`,
         TS_PCODE = `Township Pcode`, TS = `Township Name`,
         TT_HOUSEHOLDS = `Total households`, RADIO = Radio,
         TV = Television, LLPHONE = `Land line phone`,
         MPHONE = `Mobile phone`, COMPUTER = Computer,
         INTERNET = `Internet at home`) %>%
  mutate(RADIO_PR = RADIO / TT_HOUSEHOLDS * 1000,
         TV_PR = TV / TT_HOUSEHOLDS * 1000,
         LLPHONE_PR = LLPHONE / TT_HOUSEHOLDS * 1000,
         MPHONE_PR = MPHONE / TT_HOUSEHOLDS * 1000,
         COMPUTER_PR = COMPUTER / TT_HOUSEHOLDS * 1000,
         INTERNET_PR = INTERNET / TT_HOUSEHOLDS * 1000)

rate_vars <- c("RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR",
               "COMPUTER_PR", "INTERNET_PR")
rate_labels <- c(RADIO_PR = "Radio", TV_PR = "Television",
                 LLPHONE_PR = "Landline phone", MPHONE_PR = "Mobile phone",
                 COMPUTER_PR = "Computer", INTERNET_PR = "Internet at home")
```

```{r}
#| label: validate-inputs
#| include: false
stopifnot(nrow(shan_sf) == 55L, nrow(ict_derived) == 55L,
          !anyDuplicated(shan_sf$TS_PCODE),
          !anyDuplicated(ict_derived$TS_PCODE),
          setequal(shan_sf$TS_PCODE, ict_derived$TS_PCODE),
          all(st_is_valid(shan_sf)), all(ict_derived$TT_HOUSEHOLDS > 0),
          all(is.finite(as.matrix(ict_derived[rate_vars]))),
          all(as.matrix(ict_derived[rate_vars]) >= 0),
          all(as.matrix(ict_derived[rate_vars]) <= 1000))
```

The data contain `r nrow(ict_derived)` townships with complete ICT indicators. Every township code matches a boundary polygon.

## Exploratory Data Analysis

### Counts and rates

The first comparison uses radio ownership. The count describes how many households own a radio, while the rate describes how common radio ownership is within a township. The boxplots show the median, interquartile range and unusually high or low observations.

```{r}
#| label: radio-distributions
#| fig-cap: "Radio ownership: household counts and rates per 1,000 households."
radio_count_hist <- ggplot(ict_derived, aes(RADIO)) +
  geom_histogram(bins = 20, fill = "lightblue", colour = "white") +
  labs(x = "Households owning a radio", y = "Townships")
radio_count_box <- ggplot(ict_derived, aes(x = RADIO, y = "")) +
  geom_boxplot(fill = "lightblue") + labs(x = "Households owning a radio", y = NULL)
radio_rate_hist <- ggplot(ict_derived, aes(RADIO_PR)) +
  geom_histogram(bins = 20, fill = "lightblue", colour = "white") +
  labs(x = "Radio ownership per 1,000 households", y = "Townships")
radio_rate_box <- ggplot(ict_derived, aes(x = RADIO_PR, y = "")) +
  geom_boxplot(fill = "lightblue") + labs(x = "Radio ownership per 1,000 households", y = NULL)
ggarrange(radio_count_hist, radio_rate_hist, radio_count_box, radio_rate_box,
          ncol = 2, nrow = 2, heights = c(2, 1))
```

### Comparing all six indicators

These histograms put the indicators on the same unit: households per 1,000. The horizontal ranges vary between panels so that the shape of each distribution remains visible.

```{r}
#| label: ict-histograms
#| fig-cap: "Distribution of the six ICT penetration rates across Shan townships."
ict_long <- ict_derived %>%
  select(TS_PCODE, all_of(rate_vars)) %>%
  pivot_longer(all_of(rate_vars), names_to = "Indicator", values_to = "Rate")
ggplot(ict_long, aes(Rate)) +
  geom_histogram(bins = 20, fill = "lightblue", colour = "white") +
  facet_wrap(~ factor(Indicator, levels = rate_vars), scales = "free_x", ncol = 3,
             labeller = as_labeller(rate_labels)) +
  labs(x = "Households per 1,000", y = "Townships")
```

### Joining the spatial and census data

```{r}
#| label: join-data
shan_sf <- left_join(
  shan_sf,
  ict_derived %>% select(TS_PCODE, TT_HOUSEHOLDS, RADIO, TV, LLPHONE,
                        MPHONE, COMPUTER, INTERNET, all_of(rate_vars)),
  by = "TS_PCODE")
dir.create("data/rds", recursive = TRUE, showWarnings = FALSE)
write_rds(shan_sf, "data/rds/shan_sf.rds")
shan_sf <- read_rds("data/rds/shan_sf.rds")
```

### Mapping the indicators

I use five Jenks classes to show variation within each variable. Each map has its own legend, so matching colours across different variables do not represent the same quantity.

```{r}
#| label: map-helper
indicator_map <- function(variable, title) {
  tm_shape(shan_sf) +
    tm_polygons(fill = variable,
                fill.scale = tm_scale_intervals(style = "jenks", n = 5,
                                                values = "brewer.blues"),
                fill.legend = tm_legend(title = title)) +
    tm_borders() + tm_title(title)
}
```

```{r}
#| label: radio-rate-map
#| fig-cap: "Radio ownership per 1,000 households, using five Jenks classes."
radio_map <- indicator_map("RADIO_PR", "Radio per 1,000 households")
radio_map
```

```{r}
#| label: count-maps
#| fig-cap: "Total households and households owning a radio."
household_map <- indicator_map("TT_HOUSEHOLDS", "Total households")
radio_count_map <- indicator_map("RADIO", "Households owning a radio")
tmap_arrange(household_map, radio_count_map, ncol = 2)
```

```{r}
#| label: count-rate-maps
#| fig-cap: "Township size compared with the prevalence of radio ownership."
tmap_arrange(household_map, radio_map, ncol = 2)
```

The rate map answers a different question from the count map. A large number of radio-owning households can reflect township size, while a high penetration rate means radio ownership is common among its households.

## Correlation Analysis

```{r}
#| label: correlation
#| fig-cap: "Pairwise correlations among the six ICT penetration rates."
cluster_vars_cor <- cor(ict_derived[rate_vars])
corrplot.mixed(cluster_vars_cor, lower = "ellipse", upper = "number",
               tl.pos = "lt", diag = "l", tl.col = "black",
               tl.cex = 0.85, number.cex = 0.85)
```

Computer ownership and home internet access have a correlation of `r round(cluster_vars_cor["COMPUTER_PR", "INTERNET_PR"], 3)`. Following the chapter, I retain computer ownership and omit internet access from the clustering inputs to reduce the repeated influence of closely related indicators.

## Hierarchical Cluster Analysis

### Selecting the clustering variables

The five inputs are radio, television, landline phone, mobile phone and computer penetration rates. Township names label the rows, and township codes retain the link to the polygons.

```{r}
#| label: clustering-inputs
features <- c("RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
shan_ict <- shan_sf %>% st_drop_geometry() %>%
  select(all_of(features)) %>% as.data.frame()
rownames(shan_ict) <- shan_sf$TS
write_rds(shan_ict, "data/rds/shan_ict.rds")
```

### Comparing standardisation methods

Min-max standardisation puts each variable between zero and one. Z-score standardisation expresses a value relative to its variable's mean and standard deviation; it does not require the original observations to be normally distributed. Neither transformation removes skewness.

```{r}
#| label: standardise-data
shan_ict_std <- heatmaply::normalize(shan_ict)
shan_ict_z <- scale(shan_ict)
standardisation <- tibble(
  `Raw rate` = shan_ict$RADIO_PR,
  `Min-max` = shan_ict_std$RADIO_PR,
  `Z-score` = shan_ict_z[, "RADIO_PR"]) %>%
  pivot_longer(everything(), names_to = "Scale", values_to = "Value") %>%
  mutate(Scale = factor(Scale, levels = c("Raw rate", "Min-max", "Z-score")))
```

```{r}
#| label: standardisation-histograms
#| fig-cap: "Radio penetration under raw, min-max and z-score scales."
ggplot(standardisation, aes(Value)) +
  geom_histogram(bins = 20, fill = "lightblue", colour = "white") +
  facet_wrap(~ Scale, scales = "free_x", nrow = 1) +
  labs(x = NULL, y = "Townships")
```

```{r}
#| label: standardisation-density
#| fig-cap: "The same radio distribution on three scales; density heights change with the measurement scale."
ggplot(standardisation, aes(Value)) +
  geom_density(fill = "lightblue", colour = "#324a5f") +
  facet_wrap(~ Scale, scales = "free", nrow = 1) +
  labs(x = NULL, y = "Density")
```

The chapter demonstrates both transformations but uses the **unscaled penetration rates** for its main clustering models. I keep that choice here. The variables share a unit, although those with larger variation can have more influence on Euclidean distances.

### Distances and the Ward hierarchy

```{r}
#| label: ward-dendrogram
#| fig-height: 7
#| fig-cap: "Hierarchical clustering of the five raw penetration rates using Euclidean distance and ward.D."
proxmat <- dist(shan_ict, method = "euclidean")
hclust_ward <- hclust(proxmat, method = "ward.D")
plot(hclust_ward, cex = 0.55, hang = -1,
     main = "Shan township ICT profiles: ward.D", xlab = "", sub = "")
```

### Comparing linkage methods

I compare the agglomerative coefficients from `agnes()`. A larger coefficient indicates a stronger clustering structure under that linkage rule; it does not by itself establish the best geographical partition.

```{r}
#| label: linkage-comparison
#| fig-cap: "Agglomerative coefficients for four linkage methods."
linkage_methods <- c("average", "single", "complete", "ward")
linkage_scores <- tibble(
  Method = linkage_methods,
  Coefficient = map_dbl(linkage_methods, ~ agnes(shan_ict, method = .x)$ac))
ggplot(linkage_scores, aes(reorder(Method, Coefficient), Coefficient)) +
  geom_col(fill = "#4c8da8", width = 0.65) +
  geom_text(aes(label = sprintf("%.3f", Coefficient)), hjust = -0.15) +
  coord_flip() + scale_y_continuous(limits = c(0, 1.04)) +
  labs(x = NULL, y = "Agglomerative coefficient")
```

The largest coefficient is `r round(max(linkage_scores$Coefficient), 3)` for `r linkage_scores$Method[which.max(linkage_scores$Coefficient)]` linkage.

### Gap statistic and the six-group comparison

I compute the gap statistic for up to ten groups using 50 reference simulations and seed 12345. As in the chapter, this diagnostic uses `hcut()` with its `ward.D2` default, while the earlier example uses `ward.D`.

```{r}
#| label: gap-statistic
#| fig-cap: "Gap statistic across candidate numbers of clusters, with simulation standard errors."
set.seed(12345)
gap_stat <- clusGap(shan_ict, FUN = hcut, hc_method = "ward.D2",
                    nstart = 25, K.max = 10, B = 50)
gap_firstmax <- maxSE(gap_stat$Tab[, "gap"], gap_stat$Tab[, "SE.sim"],
                      method = "firstmax")
fviz_gap_stat(gap_stat, maxSE = list(method = "firstmax"))
```

The first-local-maximum rule selects a cluster count of **`r gap_firstmax`**. I retain **six groups** for the following maps, as the chapter does, to compare the clustering approaches. Six is therefore a comparison setting, rather than the formal optimum from this rule.

```{r}
#| label: ward-six-groups
#| fig-height: 7
#| fig-cap: "The ward.D hierarchy cut into six groups for comparison."
cluster_colours <- c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", "#e6ab02")
plot(hclust_ward, cex = 0.55, hang = -1,
     main = "Six groups from the ward.D hierarchy", xlab = "", sub = "")
rect.hclust(hclust_ward, k = 6, border = cluster_colours)
```

### Interactive cluster heatmap

The heatmap uses min-max scaled rates so that differences within each indicator are easier to see. Its row hierarchy is fitted to these scaled values, so its six row groups need not match the raw-rate hierarchy shown above. Hover over a cell to inspect it. On a narrow screen, scroll horizontally to see the full chart.

:::: {style="overflow-x: auto;"}
::: {style="min-width: 750px;"}

```{r}
#| label: interactive-heatmap
#| results: asis
#| fig-height: 9
heatmaply(as.matrix(shan_ict_std), Colv = NA,
          height = 950,
          dist_method = "euclidean", hclust_method = "ward.D",
          seriate = "OLO", k_row = 6,
          colors = colorRampPalette(c("#f7fbff", "#08306b")),
          margins = c(65, 150, 55, 15), fontsize_row = 9, fontsize_col = 11,
          main = "Shan ICT profiles: min-max scaled rates",
          xlab = "ICT indicators", ylab = "Townships")
```

:::
::::

### Mapping the hierarchy

```{r}
#| label: hierarchical-map
#| fig-cap: "Six clusters from the raw-rate ward.D hierarchy."
shan_sf_cluster <- shan_sf %>%
  mutate(CLUSTER = factor(cutree(hclust_ward, k = 6)))
cluster_map <- function(data, variable, title) {
  tm_shape(data) +
    tm_polygons(fill = variable,
                fill.scale = tm_scale_categorical(values = cluster_colours),
                fill.legend = tm_legend(title = "Cluster")) +
    tm_borders() + tm_title(title)
}
hierarchical_map <- cluster_map(shan_sf_cluster, "CLUSTER", "Hierarchical clusters")
hierarchical_map
```

This model uses ICT similarity without a spatial constraint. Townships with similar profiles can therefore share a group even when they are separated geographically.

## Spatially Constrained Clustering with SKATER

### Building the neighbour graph

SKATER starts with a graph of neighbouring townships. I use Queen contiguity, which accepts either a shared boundary edge or a shared vertex.

```{r}
#| label: queen-neighbour-map
#| fig-cap: "Queen contiguity links between Shan townships."
shan_sp <- as_Spatial(shan_sf)
shan_nb <- poly2nb(shan_sp, queen = TRUE)
coords <- st_coordinates(st_centroid(st_geometry(shan_sf)))
plot(st_geometry(shan_sf), border = "grey55")
plot(shan_nb, coords, col = "#246a9b", add = TRUE)
```

The graph has `r length(shan_nb)` townships and `r sum(card(shan_nb))` directed neighbour links. All townships belong to one connected component, allowing a spanning tree to cover the whole study area.

### Edge costs and the minimum spanning tree

Each neighbour link receives a cost based on the difference between the two townships' five ICT rates. The minimum spanning tree connects all townships while minimising the total edge cost. I retain the original costs with `style = "B"`.

```{r}
#| label: minimum-spanning-tree
#| fig-cap: "Minimum spanning tree based on ICT differences between adjacent townships."
lcosts <- nbcosts(shan_nb, shan_ict)
shan_w <- nb2listw(shan_nb, glist = lcosts, style = "B")
shan_mst <- mstree(shan_w, ini = 1)
plot(st_geometry(shan_sf), border = "grey55")
segments(coords[shan_mst[, 1], 1], coords[shan_mst[, 1], 2],
         coords[shan_mst[, 2], 1], coords[shan_mst[, 2], 2],
         col = "#246a9b", lwd = 1.2)
points(coords, pch = 16, cex = 0.45, col = "#246a9b")
```

### Cutting the tree into six regions

Five cuts produce six groups. Since SKATER removes edges from the neighbour tree, each resulting group remains connected through the original Queen graph.

```{r}
#| label: skater-tree
#| fig-cap: "The minimum spanning tree after SKATER divides it into six connected groups."
clust6 <- spdep::skater(edges = shan_mst[, 1:2], data = shan_ict,
                       method = "euclidean", ncuts = 5)
plot(st_geometry(shan_sf), border = "grey55")
kept_edges <- shan_mst[
  clust6$groups[shan_mst[, 1]] == clust6$groups[shan_mst[, 2]], 1:2]
segments(coords[kept_edges[, 1], 1], coords[kept_edges[, 1], 2],
         coords[kept_edges[, 2], 1], coords[kept_edges[, 2], 2],
         col = cluster_colours[clust6$groups[kept_edges[, 1]]], lwd = 1.5)
points(coords, pch = 16, cex = 0.5, col = cluster_colours[clust6$groups])
```

```{r}
#| label: skater-map
#| fig-cap: "SKATER's six spatially connected ICT regions."
shan_sf_spatialcluster <- shan_sf %>%
  mutate(SP_CLUSTER = factor(clust6$groups))
skater_map <- cluster_map(shan_sf_spatialcluster, "SP_CLUSTER", "SKATER regions")
skater_map
```

```{r}
#| label: hierarchical-skater-comparison
#| fig-cap: "Raw-rate hierarchical clustering and SKATER. Cluster numbers are separate labels within each method."
tmap_arrange(hierarchical_map, skater_map, ncol = 2)
```

The SKATER groups contain between `r min(table(clust6$groups))` and `r max(table(clust6$groups))` townships. The spatial constraint creates connected regions, but it does not require equally sized groups. A cluster number or colour in one method does not identify the same group in another.

## Spatially Constrained Clustering with ClustGeo

### Ward-like clustering without geography

ClustGeo first provides a Ward-like hierarchy based on the ICT distance matrix alone. This gives a baseline before geographical distance is introduced.

```{r}
#| label: clustgeo-baseline-dendrogram
#| fig-height: 7
#| fig-cap: "ClustGeo's unconstrained hierarchy, cut into six groups."
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.55, hang = -1,
     main = "ClustGeo baseline: ICT distance only", xlab = "", sub = "")
rect.hclust(nongeo_cluster, k = 6, border = cluster_colours)
```

```{r}
#| label: clustgeo-baseline-map
#| fig-cap: "The six ClustGeo baseline groups before adding geographical distance."
shan_sf_ngeo_cluster <- shan_sf %>%
  mutate(CLUSTER = factor(cutree(nongeo_cluster, k = 6)))
nongeo_map <- cluster_map(shan_sf_ngeo_cluster, "CLUSTER", "ClustGeo: ICT only")
nongeo_map
```

### Balancing ICT and geographical similarity

ClustGeo combines the ICT distance matrix with distances between township polygons. Touching polygons have zero separation in the latter matrix. The mixing parameter $\alpha$ controls the influence of geography: zero uses ICT distance alone, while one uses geographical distance alone.

```{r}
#| label: alpha-choice
#| fig-height: 6
#| fig-cap: "ClustGeo diagnostics across alpha values from zero to one."
geographic_distances <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(units::drop_units(geographic_distances))
alpha_diagnostics <- choicealpha(proxmat, distmat,
                                 range.alpha = seq(0, 1, 0.1), K = 6,
                                 graph = TRUE)
```

Following the chapter, I choose $\alpha=0.2$ as a compromise that adds geographical cohesion while retaining emphasis on ICT similarity. This is a soft distance constraint: unlike SKATER, it does not guarantee that every group is connected through shared boundaries.

```{r}
#| label: clustgeo-spatial-map
#| fig-cap: "ClustGeo with alpha = 0.2 and six groups."
clustG <- hclustgeo(proxmat, distmat, alpha = 0.2)
shan_sf_Gcluster <- shan_sf %>%
  mutate(CLUSTER = factor(cutree(clustG, k = 6)))
cluster_map(shan_sf_Gcluster, "CLUSTER", "ClustGeo: alpha = 0.2")
```

## Visual Interpretation of Clusters

### Radio ownership within groups

The chapter profiles the **ClustGeo baseline groups**, so I use that same partition here. Each boxplot shows within-group variation in radio penetration, rather than only one representative value per group.

```{r}
#| label: cluster-radio-boxplots
#| fig-cap: "Radio penetration in the six ClustGeo baseline groups."
ggplot(shan_sf_ngeo_cluster, aes(CLUSTER, RADIO_PR, fill = CLUSTER)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = cluster_colours) +
  labs(x = "Baseline cluster", y = "Radio-owning households per 1,000")
```

### Multivariate profiles

The parallel-coordinate plots show how the five rates vary together within each baseline group. The chapter's `globalminmax` option keeps the original rates and gives all five axes a common range. The interactive heatmap, in comparison, scales each variable separately between zero and one.

```{r}
#| label: parallel-coordinates
#| fig-height: 8
#| fig-cap: "Five-variable ICT profiles within each ClustGeo baseline cluster."
profile_data <- st_drop_geometry(shan_sf_ngeo_cluster) %>%
  select(CLUSTER, all_of(features)) %>% as.data.frame()
ggparcoord(profile_data, columns = match(features, names(profile_data)),
           groupColumn = 1, scale = "globalminmax", alphaLines = 0.2,
           boxplot = TRUE, showPoints = FALSE) +
  facet_wrap(~ CLUSTER, ncol = 3) +
  scale_colour_manual(values = cluster_colours, guide = "none") +
  scale_x_discrete(labels = rate_labels[features]) +
  labs(x = NULL, y = "Households per 1,000 (common scale)") +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

Finally, I plot the group means in the original units. Separate vertical scales make differences in the less common forms of access visible without changing their reported values.

```{r}
#| label: cluster-mean-profiles
#| fig-height: 7
#| fig-cap: "Mean ICT penetration rates for each ClustGeo baseline group."
cluster_means <- profile_data %>%
  group_by(CLUSTER) %>% summarise(across(all_of(features), mean), .groups = "drop")
cluster_means %>%
  pivot_longer(all_of(features), names_to = "Indicator", values_to = "Mean") %>%
  ggplot(aes(CLUSTER, Mean, colour = CLUSTER)) +
  geom_point(size = 3) +
  facet_wrap(~ factor(Indicator, levels = features), ncol = 3, scales = "free_y",
             labeller = as_labeller(rate_labels)) +
  scale_colour_manual(values = cluster_colours, guide = "none") +
  labs(x = "Baseline cluster", y = "Mean households per 1,000")
```

Baseline group `r as.character(cluster_means$CLUSTER[which.max(cluster_means$TV_PR)])` has the highest average television, mobile-phone and computer penetration. Radio ownership is highest in group `r as.character(cluster_means$CLUSTER[which.max(cluster_means$RADIO_PR)])`, so the groups do not follow the same ranking for every form of ICT access.

The three clustering approaches answer related questions: the hierarchical models group ICT profiles, SKATER requires connected regions, and ClustGeo allows a gradual trade-off between ICT similarity and geographical cohesion. The results describe the supplied 2014 census indicators and should not be read as current ICT access levels.

```{r}
#| label: validate-results
#| include: false
connected_group <- function(ids, nb) {
  reached <- ids[1]
  repeat {
    expanded <- sort(intersect(ids, unique(c(reached, unlist(nb[reached])))))
    if (identical(sort(reached), expanded)) break
    reached <- expanded
  }
  length(reached) == length(ids)
}
stopifnot(nrow(shan_sf) == 55L, !anyNA(shan_sf[features]),
          identical(rownames(shan_ict), shan_sf$TS),
          sum(card(shan_nb)) == 264L, n.comp.nb(shan_nb)$nc == 1L,
          nrow(shan_mst) == 54L,
          length(unique(clust6$groups)) == 6L,
          all(vapply(split(seq_len(nrow(shan_sf)), clust6$groups),
                     connected_group, logical(1), nb = shan_nb)),
          all(is.finite(as.numeric(proxmat))),
          all(is.finite(as.numeric(distmat))))
write_rds(list(
  township_count = nrow(shan_sf), correlation = cluster_vars_cor,
  linkage = linkage_scores, gap_firstmax = gap_firstmax,
  gap = gap_stat$Tab, skater_sizes = table(clust6$groups),
  baseline_means = cluster_means,
  partitions = st_drop_geometry(shan_sf) %>% select(TS_PCODE, TS) %>%
    mutate(ward = shan_sf_cluster$CLUSTER,
           skater = shan_sf_spatialcluster$SP_CLUSTER,
           clustgeo_baseline = shan_sf_ngeo_cluster$CLUSTER,
           clustgeo_alpha_02 = shan_sf_Gcluster$CLUSTER)),
  "data/rds/analysis_results.rds")
```

## References

- Kam, T. S. [Geographical Segmentation with Spatially Constrained Clustering Techniques](https://r4gdsa.netlify.app/chap12.html), *R for Geospatial Data Science and Analytics*.
- [The 2014 Myanmar Population and Housing Census](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet). The exercise uses the supplied Shan ICT extract and township boundaries.
- [Introduction to ClustGeo](https://cran.r-project.org/web/packages/ClustGeo/vignettes/intro_ClustGeo.html).
- **GGally** documentation: [parallel coordinate plots and scaling options](https://ggobi.github.io/ggally/reference/ggparcoord.html).
