---
title: "Hands-on Exercise 23: Analytical Mapping"
author: "Bao Sihan"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: true
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

## 23.1 Overview

In this exercise, I continue with choropleth mapping but move beyond standard count maps. The main idea is that raw counts can be misleading when the underlying units differ a lot in size. Therefore, I focus on analytical maps such as rate maps, percentile maps and box maps.

## 23.2 Getting Started

### 23.2.1 Installing and loading packages

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

### 23.2.2 Importing data

```{r}
NGA_wp <- read_rds("../data/NGA_wp.rds")
```

```{r}
NGA_wp
```

## 23.3 Basic Choropleth Mapping

### 23.3.1 Visualising distribution of total and functional water points

```{r}
p1 <- tm_shape(NGA_wp) +
  tm_fill("total_wp", style = "quantile", palette = "Blues") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Distribution of total water point by LGAs")

p2 <- tm_shape(NGA_wp) +
  tm_fill("wp_functional", style = "quantile", palette = "Blues") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Distribution of functional water point by LGAs")

tmap_arrange(p1, p2, nrow = 1)
```

**Observation:** The two maps look similar in some regions because areas with many total water points also tend to have many functional water points. However, count maps alone do not show how efficient or reliable the water-point system is.

## 23.4 Choropleth Map for Rates

### 23.4.1 Deriving proportion of functional and non-functional water points

```{r}
NGA_wp <- NGA_wp %>%
  mutate(
    pct_functional = wp_functional / total_wp,
    pct_nonfunctional = wp_nonfunctional / total_wp
  )
```

### 23.4.2 Plotting rate map

```{r}
tm_shape(NGA_wp) +
  tm_fill("pct_functional", style = "fixed",
          breaks = seq(0, 1, by = 0.1),
          palette = "Blues") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Rate map of functional water point by LGAs")
```

**Insight:** This map is more meaningful than the count map because it standardises by total water points. Regions with fewer total points are no longer automatically shown as unimportant.

## 23.5 Extreme Value Maps

### 23.5.1 Percentile map

#### 23.5.1.1 Data preparation

```{r}
NGA_wp <- NGA_wp %>% drop_na()
```

```{r}
percent <- c(0, 0.01, 0.1, 0.5, 0.9, 0.99, 1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)

quantile(var[, 1], percent)
```

**Student note:** When variables are extracted from an `sf` object, the geometry comes along as well. That is why `st_set_geometry(NULL)` is necessary here before using base functions like `quantile()`.

#### 23.5.1.2 Creating `get.var()` function

```{r}
get.var <- function(vname, df) {
  v <- df[vname] %>%
    st_set_geometry(NULL)
  v <- unname(v[, 1])
  return(v)
}
```

#### 23.5.1.3 Percentile mapping function

```{r}
percentmap <- function(vnam, df, legtitle = NA, mtitle = "Percentile Map") {
  percent <- c(0, 0.01, 0.1, 0.5, 0.9, 0.99, 1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)

  tm_shape(df) +
    tm_polygons() +
    tm_shape(df) +
    tm_fill(
      vnam,
      title = legtitle,
      breaks = bperc,
      palette = "Blues",
      labels = c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%")
    ) +
    tm_borders() +
    tm_layout(
      main.title = mtitle,
      legend.position = c("right", "bottom")
    )
}
```

#### 23.5.1.4 Test the percentile mapping function

```{r}
percentmap("total_wp", NGA_wp)
```

**Observation:** The percentile map highlights the extreme ends of the distribution more clearly than a normal quantile map. It is useful when I specifically want to identify unusually low or unusually high areas.

### 23.5.2 Box map

First, I use a boxplot to understand the distribution of the non-functional water points.

```{r}
ggplot(data = NGA_wp,
       aes(x = "", y = wp_nonfunctional)) +
  geom_boxplot()
```

**Observation:** There are clear upper outliers, so a box map can help carry this boxplot logic into space.

#### 23.5.2.1 Creating `boxbreaks()` function

```{r}
boxbreaks <- function(v, mult = 1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr

  bb <- vector(mode = "numeric", length = 7)

  if (lofence < qv[1]) {
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }

  if (upfence > qv[5]) {
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }

  bb[3:5] <- qv[2:4]
  return(bb)
}
```

#### 23.5.2.2 Reusing `get.var()` and testing the breaks

```{r}
var <- get.var("wp_nonfunctional", NGA_wp)
boxbreaks(var)
```

#### 23.5.2.3 Box map function

```{r}
boxmap <- function(vnam, df, legtitle = NA, mtitle = "Box Map", mult = 1.5) {
  var <- get.var(vnam, df)
  bb <- boxbreaks(var)

  tm_shape(df) +
    tm_polygons() +
    tm_shape(df) +
    tm_fill(
      vnam,
      title = legtitle,
      breaks = bb,
      palette = "Blues",
      labels = c("lower outlier",
                 "< 25%",
                 "25% - 50%",
                 "50% - 75%",
                 "> 75%",
                 "upper outlier")
    ) +
    tm_borders() +
    tm_layout(
      main.title = mtitle,
      title.position = c("left", "top")
    )
}
```

#### 23.5.2.4 Test the box map function

```{r}
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```

**Insight:** The box map is useful because it keeps the quartile structure while also flagging extreme observations explicitly. In my view, this makes it more interpretable than a standard choropleth when the purpose is outlier detection.

## 23.6 Final reflection

This chapter helped me understand why analytical mapping is not only about aesthetics. It is mainly about choosing a mapping logic that matches the question. If I care about relative performance, I should use rates. If I care about outliers, percentile maps and box maps are more informative than standard classed choropleths.
