---
title: "Hands-on Exercise 14: Heatmap for Visualising and Analysing Multivariate Data"
author: "Bao Sihan"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  warning: false
  message: false
---

## 14.1 Overview

Heatmaps show multivariate patterns using color intensity (good for spotting **clusters**, **similarity**, and **outliers**).\
In this hands-on, we will create:

-   static heatmaps (base `heatmap()`),
-   interactive heatmaps using **heatmaply**,
-   different data transformations (scale / normalize / percentize),
-   clustering method selection + number of clusters,
-   ordering with **seriation**.

## 14.2 Installing and launching R packages

```{r}
pacman::p_load(seriation, dendextend, heatmaply, tidyverse)
```

## 14.3 Importing and preparing the data set

We use World Happiness 2018 dataset (CSV in `data/` folder).

```{r}
wh <- read_csv("../data/WHData-2018.csv")
glimpse(wh)
```

### 14.3.1 Use Country as row names

```{r}
wh <- wh %>% mutate(Country = as.character(Country))
row.names(wh) <- wh$Country
```

### 14.3.2 Transform data frame into matrix

Select columns that represent numeric indicators (adjust indices if your file differs).

```{r}
wh1 <- wh %>% dplyr::select(c(3, 7:12))
wh_matrix <- data.matrix(wh1)
```

## 14.4 Static heatmap (base heatmap())

### 14.4.1 Default heatmap

```{r}
wh_heatmap <- heatmap(wh_matrix, Rowv = NA, Colv = NA)
```

### 14.4.2 Cluster heatmap + scaling

Because variables have different ranges, scaling often helps.

```{r}
wh_heatmap <- heatmap(
  wh_matrix,
  scale = "column",
  cexRow = 0.6,
  cexCol = 0.8,
  margins = c(10, 4)
)
```

## 14.5 Creating interactive heatmap (heatmaply)

### 14.5.1 Basic heatmaply

```{r}
heatmaply(wh_matrix)
```

### 14.5.2 Data transformation options

**Scaling** (z-score) column-wise:

```{r}
heatmaply(wh_matrix, scale = "column")
```

**Normalising** (0–1 scale) on the input data (typically after selecting numeric cols):

```{r}
heatmaply(normalize(wh_matrix))
```

**Percentising** (convert each value into percentile rank):

```{r}
heatmaply(percentize(wh_matrix))
```

### 14.5.3 Clustering method & number of clusters

You can compare clustering methods using `dendextend` (one practical workflow is to compute distances and evaluate).\
Below shows a commonly-used approach consistent with the lecture workflow: estimate the number of clusters (k) via silhouette.

```{r}
# distance and clustering
wh_d <- dist(normalize(wh_matrix), method = "euclidean")
wh_clust <- hclust(wh_d, method = "average")

# find a reasonable k using silhouette (heatmaply provides helpers in some setups)
num_k <- find_k(wh_clust)
plot(num_k)
```

Then plot with a selected k (example: k = 3):

```{r}
heatmaply(
  normalize(wh_matrix),
  dist_method = "euclidean",
  hclust_method = "average",
  k_row = 3
)
```

### 14.5.4 Seriation (re-ordering)

Seriation reorders rows/columns to make similar items adjacent. Try different methods:

```{r}
heatmaply(normalize(wh_matrix), seriate = "OLO")
heatmaply(normalize(wh_matrix), seriate = "GW")
heatmaply(normalize(wh_matrix), seriate = "mean")
heatmaply(normalize(wh_matrix), seriate = "none")
```

### 14.5.5 Colour palettes

Use a single-hue palette (e.g., Blues) to avoid misleading diverging colors when all values are positive.

```{r}
heatmaply(
  normalize(wh_matrix),
  seriate = "none",
  colors = viridisLite::viridis(256)
)
```

### 14.5.6 Finishing touch

Add clearer labels and tune margins/font sizes.

```{r}
heatmaply(
  normalize(wh_matrix),
  seriate = "none",
  colors = viridisLite::viridis(256),
  k_row = 5,
  margins = c(50, 200, 60, 20),
  fontsize_row = 4,
  fontsize_col = 5,
  main = "World Happiness Score and Variables by Country, 2018",
  xlab = "World Happiness Indicators",
  ylab = "World Countries"
)
```

## Notes / Tips

-   Heatmaps are very sensitive to scaling: always decide if you need raw vs scaled values.
-   If your row labels become unreadable, consider filtering top-N countries or using interactive mode.
