---
title: "Hands-on Exercise 2B: Second-order Spatial Point Patterns Analysis"
author: "Bao Sihan"
date: "September 1, 2026"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Overview

In this hands-on exercise, I use second-order spatial point pattern analysis to examine how childcare centre locations are related to one another at different distances. Unlike first-order analysis, which focuses on changes in point intensity, second-order analysis evaluates whether the observed points show clustering, dispersion, or randomness across a range of spatial scales.

I apply the G-, F-, K-, and L-functions to childcare centres in Choa Chu Kang and Tampines. Monte Carlo simulation envelopes are used to compare each observed pattern with complete spatial randomness (CSR).

## The Data

I use the two data sets specified in the exercise:

- **Child Care Services** from [data.gov.sg](https://data.gov.sg/datasets/d_5d668e3f544335f8028f546827b773b4/view); and
- **Master Plan 2019 Subzone Boundary (No Sea)** from [data.gov.sg](https://data.gov.sg/datasets/d_8594ae9ff96d0c708bc2af633048edfb/view).

The GeoJSON files are stored in this exercise's `data/geospatial` folder so that the page can be rendered independently with relative paths.

## Installing and Loading the R Packages

I use **sf** for geospatial data, **spatstat** for the second-order analysis, **tmap** as listed in the exercise, and **tidyverse** for data preparation.

```{r}
#| label: load-packages
pacman::p_load(sf, spatstat, tmap, tidyverse)
```

## Data Import and Preparation

The preparation follows the same process used in Hands-on Exercise 2A. I import both layers, transform them to Singapore SVY21 (EPSG:3414), remove the specified offshore planning areas, and convert the childcare locations into a planar point pattern.

```{r}
#| label: import-and-prepare-data
mpsz_cl <- st_read(
  "data/geospatial/MasterPlan2019SubzoneBoundaryNoSea.geojson"
) %>%
  st_transform(crs = 3414) %>%
  st_make_valid() %>%
  filter(
    SUBZONE_N != "SOUTHERN GROUP",
    PLN_AREA_N != "WESTERN ISLANDS",
    PLN_AREA_N != "NORTH-EASTERN ISLANDS"
  ) %>%
  select(
    REGION_N,
    PLN_AREA_N,
    SUBZONE_N,
    SUBZONE_C,
    geometry
  )

childcare_sf <- st_read(
  "data/geospatial/ChildCareServices.geojson"
) %>%
  st_transform(crs = 3414)

childcare_ppp <- as.ppp(childcare_sf)
```

I extract Choa Chu Kang and Tampines, convert their boundaries into observation windows, and clip the childcare point pattern to each window.

```{r}
#| label: prepare-study-areas
ck <- mpsz_cl %>%
  filter(PLN_AREA_N == "CHOA CHU KANG")

tm <- mpsz_cl %>%
  filter(PLN_AREA_N == "TAMPINES")

ck_owin <- as.owin(ck)
tm_owin <- as.owin(tm)

childcare_ck_ppp <- childcare_ppp[ck_owin]
childcare_tm_ppp <- childcare_ppp[tm_owin]

childcare_ck_ppp
childcare_tm_ppp
```

There are `r npoints(childcare_ck_ppp)` childcare centre points in the Choa Chu Kang observation window and `r npoints(childcare_tm_ppp)` in the Tampines observation window.

I set a random seed so that the simulation envelopes can be reproduced.

```{r}
#| label: set-random-seed
set.seed(1234)
```

## Second-order Spatial Point Patterns Analysis

For each function, I use the following hypotheses:

- $H_0$: Childcare centres are randomly distributed in the planning area.
- $H_1$: Childcare centres are not randomly distributed in the planning area.

The observed function is compared with the CSR simulation envelope. An observed curve outside the envelope indicates a departure from spatial randomness over the corresponding distance range.

## Analysing the Spatial Point Process Using the G-function

The G-function measures the distribution of nearest-neighbour distances from each observed childcare centre to the next closest centre. A curve above the CSR expectation at short distances indicates that nearest neighbours occur more quickly than expected, which is consistent with clustering.

### Choa Chu Kang

#### Computing the G-function estimate

```{r}
#| label: g-function-ck
#| fig-width: 8
#| fig-height: 6
G_CK <- Gest(
  childcare_ck_ppp,
  correction = "border"
)

plot(
  G_CK,
  xlim = c(0, 500),
  main = "G-function for Choa Chu Kang"
)
```

#### Performing the CSR test

I generate a Monte Carlo envelope using 999 CSR simulations.

```{r}
#| label: g-envelope-ck
#| fig-width: 8
#| fig-height: 6
G_CK_csr <- envelope(
  childcare_ck_ppp,
  Gest,
  nsim = 999,
  verbose = FALSE
)

plot(
  G_CK_csr,
  main = "CSR envelope for the Choa Chu Kang G-function"
)
```

The observed G-function rises outside the CSR envelope over part of the short-distance range. I therefore reject the null hypothesis over those distances. Childcare centres in Choa Chu Kang have shorter nearest-neighbour distances than expected under CSR, indicating local clustering.

### Tampines

#### Computing the G-function estimate

```{r}
#| label: g-function-tampines
#| fig-width: 8
#| fig-height: 6
G_tm <- Gest(
  childcare_tm_ppp,
  correction = "best"
)

plot(
  G_tm,
  main = "G-function for Tampines"
)
```

#### Performing the CSR test

```{r}
#| label: g-envelope-tampines
#| fig-width: 8
#| fig-height: 6
G_tm_csr <- envelope(
  childcare_tm_ppp,
  Gest,
  correction = "all",
  nsim = 999,
  verbose = FALSE
)

plot(
  G_tm_csr,
  main = "CSR envelope for the Tampines G-function"
)
```

The observed Tampines G-function also leaves the CSR envelope at short distances. I reject the null hypothesis over this range and conclude that the childcare centre pattern has a clustered nearest-neighbour structure.

## Analysing the Spatial Point Process Using the F-function

The F-function measures distances from arbitrary locations in the observation window to the nearest childcare centre. For a clustered pattern, larger empty spaces remain between groups of points, so the observed F-function is expected to be lower than CSR over part of the distance range.

### Choa Chu Kang

#### Computing the F-function estimate

```{r}
#| label: f-function-ck
#| fig-width: 8
#| fig-height: 6
F_CK <- Fest(childcare_ck_ppp)

plot(
  F_CK,
  main = "F-function for Choa Chu Kang"
)
```

#### Performing the CSR test

```{r}
#| label: f-envelope-ck
#| fig-width: 8
#| fig-height: 6
F_CK_csr <- envelope(
  childcare_ck_ppp,
  Fest,
  nsim = 999,
  verbose = FALSE
)

plot(
  F_CK_csr,
  main = "CSR envelope for the Choa Chu Kang F-function"
)
```

The observed F-function remains within the CSR envelope across the plotted distance range. I therefore fail to reject the null hypothesis for this test. Based on empty-space distances alone, the Choa Chu Kang F-function does not provide sufficient evidence of a departure from CSR.

### Tampines

#### Computing the F-function estimate

```{r}
#| label: f-function-tampines
#| fig-width: 8
#| fig-height: 6
F_tm <- Fest(
  childcare_tm_ppp,
  correction = "best"
)

plot(
  F_tm,
  main = "F-function for Tampines"
)
```

#### Performing the CSR test

```{r}
#| label: f-envelope-tampines
#| fig-width: 8
#| fig-height: 6
F_tm_csr <- envelope(
  childcare_tm_ppp,
  Fest,
  correction = "all",
  nsim = 999,
  verbose = FALSE
)

plot(
  F_tm_csr,
  main = "CSR envelope for the Tampines F-function"
)
```

The observed F-function for Tampines departs from the CSR envelope. I reject the null hypothesis over the affected distances. The lower observed F-values indicate empty spaces between groups of childcare centres and support a clustered pattern.

## Analysing the Spatial Point Process Using the K-function

The K-function evaluates the expected number of further events found within a given distance of an event. Following the professor's workflow, I plot $K(d)-d$ against distance. Values above the CSR envelope indicate more neighbouring centres within that distance than expected under randomness.

### Choa Chu Kang

#### Computing the K-function estimate

```{r}
#| label: k-function-ck
#| fig-width: 8
#| fig-height: 6
K_ck <- Kest(
  childcare_ck_ppp,
  correction = "Ripley"
)

plot(
  K_ck,
  . - r ~ r,
  ylab = "K(d)-r",
  xlab = "d (m)",
  main = "K-function for Choa Chu Kang"
)
```

#### Performing the CSR test

```{r}
#| label: k-envelope-ck
#| fig-width: 8
#| fig-height: 6
K_ck_csr <- envelope(
  childcare_ck_ppp,
  Kest,
  nsim = 99,
  rank = 1,
  glocal = TRUE,
  verbose = FALSE
)

plot(
  K_ck_csr,
  . - r ~ r,
  xlab = "d (m)",
  ylab = "K(d)-r",
  main = "CSR envelope for the Choa Chu Kang K-function"
)
```

The observed K-function lies above the CSR envelope over part of the distance range. I reject the null hypothesis over those distances because there are more neighbouring childcare centres than expected under CSR. This indicates clustering in Choa Chu Kang.

### Tampines

#### Computing the K-function estimate

```{r}
#| label: k-function-tampines
#| fig-width: 8
#| fig-height: 6
K_tm <- Kest(
  childcare_tm_ppp,
  correction = "Ripley"
)

plot(
  K_tm,
  . - r ~ r,
  ylab = "K(d)-r",
  xlab = "d (m)",
  xlim = c(0, 1000),
  main = "K-function for Tampines"
)
```

#### Performing the CSR test

```{r}
#| label: k-envelope-tampines
#| fig-width: 8
#| fig-height: 6
K_tm_csr <- envelope(
  childcare_tm_ppp,
  Kest,
  nsim = 99,
  rank = 1,
  glocal = TRUE,
  verbose = FALSE
)

plot(
  K_tm_csr,
  . - r ~ r,
  xlab = "d (m)",
  ylab = "K(d)-r",
  xlim = c(0, 500),
  main = "CSR envelope for the Tampines K-function"
)
```

The Tampines K-function exceeds the CSR envelope over part of the plotted distance range. I reject the null hypothesis over those distances and conclude that childcare centres are clustered at those spatial scales.

## Analysing the Spatial Point Process Using the L-function

The L-function is a transformed version of the K-function. Following the exercise, I plot $L(d)-d$ so that the CSR reference is easier to compare with the observed pattern. Positive departures above the CSR envelope indicate clustering.

### Choa Chu Kang

#### Computing the L-function estimate

```{r}
#| label: l-function-ck
#| fig-width: 8
#| fig-height: 6
L_ck <- Lest(
  childcare_ck_ppp,
  correction = "Ripley"
)

plot(
  L_ck,
  . - r ~ r,
  ylab = "L(d)-r",
  xlab = "d (m)",
  main = "L-function for Choa Chu Kang"
)
```

#### Performing the CSR test

```{r}
#| label: l-envelope-ck
#| fig-width: 8
#| fig-height: 6
L_ck_csr <- envelope(
  childcare_ck_ppp,
  Lest,
  nsim = 99,
  rank = 1,
  glocal = TRUE,
  verbose = FALSE
)

plot(
  L_ck_csr,
  . - r ~ r,
  xlab = "d (m)",
  ylab = "L(d)-r",
  main = "CSR envelope for the Choa Chu Kang L-function"
)
```

The observed L-function is above the CSR envelope over part of the distance range. I reject the null hypothesis over those distances and interpret the positive departure as evidence of clustering in Choa Chu Kang.

### Tampines

#### Computing the L-function estimate

```{r}
#| label: l-function-tampines
#| fig-width: 8
#| fig-height: 6
L_tm <- Lest(
  childcare_tm_ppp,
  correction = "Ripley"
)

plot(
  L_tm,
  . - r ~ r,
  ylab = "L(d)-r",
  xlab = "d (m)",
  xlim = c(0, 1000),
  main = "L-function for Tampines"
)
```

#### Performing the CSR test

```{r}
#| label: l-envelope-tampines
#| fig-width: 8
#| fig-height: 6
L_tm_csr <- envelope(
  childcare_tm_ppp,
  Lest,
  nsim = 99,
  rank = 1,
  glocal = TRUE,
  verbose = FALSE
)

plot(
  L_tm_csr,
  . - r ~ r,
  xlab = "d (m)",
  ylab = "L(d)-r",
  xlim = c(0, 500),
  main = "CSR envelope for the Tampines L-function"
)
```

The observed Tampines L-function rises above the CSR envelope over part of the plotted distance range. I reject the null hypothesis over those distances and conclude that the point pattern shows clustering at those spatial scales.
