---
title: "In Class Exercise 4: Geographically Weighted Summary Statistics"
author: "Bao Sihan"
date: "September 19, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
  fig-height: 6
---

## Overview

In this exercise, I use the Hunan data from Hands-on Exercise 4 to calculate geographically weighted summary statistics for GDP per capita (GDPPC). Following the in-class steps, I compare two ways of selecting an adaptive bandwidth and use the AIC-selected bandwidth to map the local mean of GDPPC.

## Getting started

I load the packages used in the exercise. The Hunan shapefile is stored in `data/geospatial`, and the 2012 data table is in `data/aspatial`. Both paths are relative to this exercise folder.

```{r}
#| label: setup
pacman::p_load(sf, ggstatsplot, tmap, tidyverse, knitr, GWmodel)
tmap_mode("plot")
```

## Importing and preparing the data

### Importing the Hunan shapefile

I read the county boundaries as an `sf` polygon object.

```{r}
#| label: import-boundaries
hunan_sf <- st_read(dsn = "data/geospatial", layer = "Hunan")
```

### Importing the Hunan 2012 table

I import the development indicators as a tibble.

```{r}
#| label: import-indicators
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### Joining the two data sets

I join the indicators to the boundaries using `County`. I select the same fields as in the class example by name, so that the selection does not depend on their column positions. The geometry is retained by `sf`.

```{r}
#| label: join-data
hunan_sf <- left_join(hunan_sf, hunan2012, by = "County") %>%
  select(NAME_2, ID_3, NAME_3, County, GDPPC, GIO, RORP, ROREmp)
```

```{r}
#| label: check-data
#| include: false
stopifnot(
  nrow(hunan_sf) == nrow(hunan2012),
  !anyDuplicated(hunan_sf$County),
  !anyNA(hunan_sf$GDPPC),
  st_is_longlat(hunan_sf)
)
```

## Mapping GDPPC

I display a labelled county map beside a GDPPC choropleth. This helps me locate the counties before calculating the local statistics.

```{r}
#| label: map-gdppc
#| fig-cap: "Hunan county names and GDP per capita in 2012."
basemap <- tm_shape(hunan_sf) +
  tm_polygons() +
  tm_text("NAME_3", size = 0.5)

gdppc <- qtm(hunan_sf, fill = "GDPPC")
tmap_arrange(basemap, gdppc, asp = 1, ncol = 2)
```

## Converting to a SpatialPolygonsDataFrame

The following `GWmodel` functions use a spatial object from the `sp` package, so I convert the joined `sf` object with `as_Spatial()`.

```{r}
#| label: convert-spatial
hunan_sp <- hunan_sf %>%
  as_Spatial()
```

## Selecting an adaptive bandwidth

I use an intercept-only model, `GDPPC ~ 1`, with a bisquare kernel. With an adaptive bandwidth, the neighbourhood size is expressed as a number of nearest neighbours rather than a fixed distance. Since the boundaries use longitude and latitude, I set `longlat = TRUE`.

### Cross-validation bandwidth

First, I select the bandwidth using cross-validation (CV).

```{r}
#| label: bandwidth-cv
bw_CV <- bw.gwr(
  GDPPC ~ 1,
  data = hunan_sp,
  approach = "CV",
  adaptive = TRUE,
  kernel = "bisquare",
  longlat = TRUE
)
bw_CV
```

### AIC bandwidth

Next, I select a bandwidth using the AIC approach. As in the class example, I use this bandwidth in the summary-statistics calculation below.

```{r}
#| label: bandwidth-aic
bw_AIC <- bw.gwr(
  GDPPC ~ 1,
  data = hunan_sp,
  approach = "AIC",
  adaptive = TRUE,
  kernel = "bisquare",
  longlat = TRUE
)
bw_AIC
```

The selected bandwidths are `r bw_CV` neighbours for CV and `r bw_AIC` neighbours for AIC.

## Computing geographically weighted summary statistics

I calculate the local summary statistics for GDPPC with `gwss()`. I keep the adaptive bisquare kernel and enable the quantile-based statistics, following the class example.

```{r}
#| label: calculate-gwss
gwstat <- gwss(
  data = hunan_sp,
  vars = "GDPPC",
  bw = bw_AIC,
  kernel = "bisquare",
  quantile = TRUE,
  adaptive = TRUE,
  longlat = TRUE
)
```

## Preparing the output data

I convert the result to a data frame and attach it to the original county polygons. The output follows the input county order. `GDPPC_LM` contains the geographically weighted local mean used in the final map.

```{r}
#| label: prepare-output
gwstat_df <- as.data.frame(gwstat$SDF)
hunan_gstat <- cbind(hunan_sf, gwstat_df)
```

```{r}
#| label: check-output
#| include: false
stopifnot(
  nrow(gwstat_df) == nrow(hunan_sf),
  all(is.finite(hunan_gstat$GDPPC_LM))
)
```

## Visualising geographically weighted summary statistics

I map the local mean using five quantile classes and a blue colour scale. Darker areas represent higher locally weighted GDPPC values. Unlike the earlier GDPPC map, each value here summarises the surrounding neighbourhood using the selected spatial weights.

```{r}
#| label: map-local-mean
#| fig-height: 9
#| echo: false
#| fig-cap: "Geographically weighted mean of GDPPC, using the AIC-selected adaptive bandwidth."
tm_shape(hunan_gstat) +
  tm_polygons(
    fill = "GDPPC_LM",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(title = "Local mean")
  ) +
  tm_borders(col_alpha = 0.5) +
  tm_title(
    "Distribution of geographically weighted mean",
    position = tm_pos_out("center", "top"),
    size = 1.2
  ) +
  tm_layout(frame = TRUE)
```
