---
title: "Hands-on Exercise 1A: Geospatial Data Science with R"
author: "Bao Sihan"
date: "August 27, 2026"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Learning Outcome

In this hands-on exercise, I learn how to use the **sf** and **tidyverse** packages to import, inspect, transform, process, and visualise geospatial data in R. I also practise converting an aspatial data frame into a simple feature data frame and applying basic geoprocessing methods.

The main tasks covered are:

- importing polygon, line, and point geospatial data;
- examining the geometry and attributes of simple feature data frames;
- assigning and transforming coordinate reference systems;
- converting longitude and latitude values into point features;
- performing buffer, intersection, and area calculations; and
- using simple exploratory plots to examine the results.

## Data Acquisition

I used the four data sets specified in the exercise:

- **Master Plan 2014 Subzone Boundary (Web)** from [data.gov.sg](https://data.gov.sg/datasets?agencies=Urban+Redevelopment+Authority+(URA)&resultId=1708);
- **Pre-Schools Location** from [data.gov.sg](https://data.gov.sg/datasets/d_61eefab99958fd70e6aab17320a71f1c/view);
- **Cycling Path** from [LTA DataMall](https://datamall.lta.gov.sg/content/datamall/en/static-data.html); and
- the latest Singapore **listings.csv** from [Inside Airbnb](https://insideairbnb.com/get-the-data/).

The geospatial files are stored in `data/geospatial`, while the Airbnb CSV file is stored in `data/aspatial`.

## Getting Started

The **sf** package is used for handling geospatial data. The **tidyverse** collection provides functions for importing, wrangling, and visualising data.

```{r}
#| label: load-packages
pacman::p_load(sf, tidyverse)
```

## Importing Geospatial Data

### Importing polygon feature data in shapefile format

I first import the Master Plan 2014 Subzone Boundary shapefile. The `dsn` argument gives the folder containing the shapefile, while `layer` gives the shapefile name without its extension.

```{r}
#| label: import-subzones
mpsz <- st_read(
  dsn = "data/geospatial",
  layer = "MP14_SUBZONE_WEB_PL"
)
```

The output shows that `mpsz` contains `r nrow(mpsz)` multipolygon features and `r ncol(mpsz) - 1` attribute fields. The layer is stored in the SVY21 projected coordinate system.

### Importing polyline feature data in shapefile format

Next, I import the Cycling Path shapefile as a line feature data frame.

```{r}
#| label: import-cycling-path
cyclingpath <- st_read(
  dsn = "data/geospatial",
  layer = "CyclingPathGazette"
)
```

The current Cycling Path data contains `r nrow(cyclingpath)` multiline features and is also stored in SVY21.

### Importing point feature data in KML format

For a KML file, I provide the complete file path and extension.

```{r}
#| label: import-preschools
preschool <- st_read(
  "data/geospatial/PreSchoolsLocation.kml"
)
```

The imported `preschool` object contains `r nrow(preschool)` point features. Unlike the first two layers, it is initially stored in the WGS 84 geographic coordinate system.

## Checking the Content of a Simple Feature Data Frame

Before carrying out any geospatial operations, I examine the geometry and attribute information in `mpsz`.

### Working with `st_geometry()`

```{r}
#| label: inspect-geometry
st_geometry(mpsz)
```

This output summarises the geometry type, bounding box, and coordinate reference system of the layer.

### Working with `glimpse()`

```{r}
#| label: inspect-attributes
glimpse(mpsz)
```

`glimpse()` allows me to see the data type and sample values for every attribute. The geometry column stores the multipolygon features.

### Working with `head()`

```{r}
#| label: inspect-first-records
head(mpsz, n = 5)
```

The `n` argument lets me control how many complete records are displayed.

## Plotting the Geospatial Data

I first use the base R `plot()` function to obtain a quick view of the subzone data.

```{r}
#| label: plot-all-subzone-attributes
#| fig-width: 10
#| fig-height: 8
plot(mpsz)
```

The default plot displays multiple maps for the attributes in the simple feature data frame.

To display only the subzone boundaries, I plot the geometry column.

```{r}
#| label: plot-subzone-geometry
#| fig-width: 8
#| fig-height: 6
plot(st_geometry(mpsz))
```

I can also plot the layer using one selected attribute.

```{r}
#| label: plot-planning-areas
#| fig-width: 9
#| fig-height: 7
plot(mpsz["PLN_AREA_N"])
```

Before transforming the pre-school layer, I try plotting it on top of the subzone layer.

```{r}
#| label: plot-before-transform
#| fig-width: 8
#| fig-height: 6
plot(st_geometry(mpsz))
plot(st_geometry(preschool), add = TRUE, col = "red", pch = 20)
```

The pre-school points do not appear in the correct position because `mpsz` uses projected SVY21 coordinates while `preschool` still uses geographic WGS 84 coordinates.

## Working with Projection

### Assigning the EPSG code to a simple feature data frame

I check the coordinate reference system of `mpsz`.

```{r}
#| label: check-subzone-crs
st_crs(mpsz)
```

Although the layers are labelled SVY21, the appropriate EPSG code for Singapore SVY21 is 3414. I assign this EPSG code to both projected layers using `st_set_crs()`.

```{r}
#| label: assign-subzone-crs
mpsz <- st_set_crs(mpsz, 3414)
cyclingpath <- st_set_crs(cyclingpath, 3414)
st_crs(mpsz)
```

### Transforming the pre-school layer from WGS 84 to SVY21

The pre-school coordinates must be mathematically transformed rather than simply relabelled. I therefore use `st_transform()`.

```{r}
#| label: transform-preschools
preschool <- st_transform(
  preschool,
  crs = 3414
)

st_geometry(preschool)
```

After the transformation, both layers use EPSG:3414. The pre-school points can now be plotted in the correct locations.

```{r}
#| label: plot-after-transform
#| fig-width: 8
#| fig-height: 6
plot(st_geometry(mpsz), col = "grey95", border = "grey60")
plot(st_geometry(preschool), add = TRUE, col = "red", pch = 20, cex = 0.25)
```

## Importing and Converting Aspatial Data

### Importing the aspatial data

The Airbnb data is stored as a CSV file. I import it as a tibble using `read_csv()`.

```{r}
#| label: import-airbnb
listings <- read_csv("data/aspatial/listings.csv")
list(listings)
```

The current file contains `r nrow(listings)` listings and `r ncol(listings)` columns. The `longitude` and `latitude` fields provide the coordinates required to create point features.

### Creating a simple feature data frame

I use `st_as_sf()` to convert the longitude and latitude columns into point geometry. The source coordinates use EPSG:4326, and I then transform them to EPSG:3414.

```{r}
#| label: convert-airbnb-to-sf
listings_sf <- st_as_sf(
  listings,
  coords = c("longitude", "latitude"),
  crs = 4326
) %>%
  st_transform(crs = 3414)

glimpse(listings_sf)
```

The new `geometry` column contains the point features, while the original longitude and latitude columns have been used to form those geometries.

I then plot the Airbnb listings on top of the planning subzones.

```{r}
#| label: plot-airbnb
#| fig-width: 8
#| fig-height: 6
plot(st_geometry(mpsz), col = "grey95", border = "grey60")
plot(st_geometry(listings_sf), add = TRUE, col = "steelblue", pch = 20, cex = 0.3)
```

## Geoprocessing with the `sf` Package

### Use case 1: Land acquisition analysis

#### The scenario

The cycling path is to be upgraded, and a five-metre strip on both sides of the path is required. I calculate the buffer area and then examine the part located within Tampines West.

#### The solution

I use `st_buffer()` to create a five-metre buffer around every cycling path feature.

```{r}
#| label: buffer-cycling-path
buffer_cycling <- st_buffer(
  cyclingpath,
  dist = 5,
  nQuadSegs = 30
)
```

Next, I calculate the area of each buffered feature.

```{r}
#| label: calculate-buffer-area
buffer_cycling <- buffer_cycling %>%
  mutate(AREA = st_area(geometry))

total_buffer_area <- sum(buffer_cycling$AREA)
total_buffer_area
```

The summed buffer area is approximately `r round(as.numeric(total_buffer_area) / 1000000, 2)` square kilometres. As in the exercise, this is the sum of the individual feature buffers.

I select Tampines West from the subzone layer and use `st_intersection()` to retain only the cycling path buffers inside it.

```{r}
#| label: select-tampines-west
mpsz_selected <- mpsz %>%
  filter(SUBZONE_N == "TAMPINES WEST")

buffer_cycling_selected <- st_intersection(
  buffer_cycling,
  mpsz_selected
)
```

```{r}
#| label: plot-tampines-west-buffer
#| fig-width: 8
#| fig-height: 6
plot(st_geometry(mpsz_selected), col = "grey95", border = "grey40")
plot(st_geometry(buffer_cycling_selected), add = TRUE, col = "lightblue", border = "steelblue")
```

The map shows the sections of the five-metre cycling path buffer that fall within Tampines West.

### Use case 2: Number of pre-schools by planning subzone

#### The scenario

The number of pre-schools in each planning subzone is required for planning. I use a spatial relationship to count the points inside every subzone and then calculate pre-school density.

#### The solution

`st_intersects()` identifies the pre-schools located inside each subzone. `lengths()` converts the resulting list into a count for every row of `mpsz`.

```{r}
#| label: count-preschools
mpsz$`PreSch Count` <- lengths(
  st_intersects(mpsz, preschool)
)

summary(mpsz$`PreSch Count`)
```

I use `top_n()` to identify the subzone with the largest number of pre-schools.

```{r}
#| label: top-preschool-subzone
top_n(mpsz, 1, `PreSch Count`)
```

Next, I calculate the area of every subzone and derive pre-school density per square kilometre.

```{r}
#| label: calculate-preschool-density
mpsz$Area <- mpsz %>%
  st_area()

mpsz <- mpsz %>%
  mutate(
    `PreSch Density` = `PreSch Count` / Area * 1000000
  )
```

I first use a base R histogram to examine the distribution.

```{r}
#| label: base-density-histogram
#| fig-width: 8
#| fig-height: 6
hist(mpsz$`PreSch Density`)
```

I then create a more clearly labelled histogram using **ggplot2**.

```{r}
#| label: ggplot-density-histogram
#| fig-width: 9
#| fig-height: 6
ggplot(
  data = mpsz,
  aes(x = as.numeric(`PreSch Density`))
) +
  geom_histogram(
    bins = 20,
    colour = "black",
    fill = "lightblue"
  ) +
  labs(
    title = "Distribution of Pre-school Density in Singapore",
    x = "Pre-school density (per square kilometre)",
    y = "Frequency"
  )
```

Finally, I use a scatterplot to compare pre-school count with pre-school density.

```{r}
#| label: preschool-scatterplot
#| fig-width: 8
#| fig-height: 6
ggplot(
  data = mpsz,
  aes(
    y = `PreSch Count`,
    x = as.numeric(`PreSch Density`)
  )
) +
  geom_point(colour = "black", fill = "lightblue") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(
    x = "Pre-school density (per square kilometre)",
    y = "Pre-school count"
  )
```

The two plots show that the distribution of pre-schools varies across planning subzones. Subzones with higher counts do not always have the highest density because their land areas are different.
