---
title: "Hands-on Exercise 1B: Thematic Mapping and GeoVisualisation with R"
author: "Bao Sihan"
date: "August 27, 2026"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Overview

In this hands-on exercise, I use thematic maps to visualise population characteristics that are not directly visible from geographic boundaries. I use **tmap** to prepare choropleth maps and to explore how map classification, colour, layout, small multiples, statistical charts, and interactive controls affect the presentation of spatial patterns.

## Getting Started

The main package used in this exercise is **tmap**. I also use **sf** for geospatial data, **readr** for importing the population file, **tidyr** for reshaping data, **dplyr** for data wrangling, and **rvest** as listed in the exercise.

```{r}
#| label: load-packages
pacman::p_load(sf, tmap, tidyverse, rvest)
```

## Importing Data into R

### The Data

Two data sets are used:

- **Master Plan 2019 Subzone Boundary (No Sea)** from [data.gov.sg](https://data.gov.sg/datasets/d_8594ae9ff96d0c708bc2af633048edfb/view). The current official release is provided in GeoJSON format and contains Singapore's planning subzone boundaries.
- **Singapore Residents by Planning Area/Subzone, Age Group, Sex and Type of Dwelling, June 2024** from the [Singapore Department of Statistics](https://www.singstat.gov.sg/). The CSV file contains the `PA` and `SZ` fields used to join the population records to the subzone boundaries.

The files are stored inside this exercise in `data/geospatial` and `data/aspatial`.

### Importing Geospatial Data into R

I use `st_read()` to import the Master Plan 2019 GeoJSON file as a simple feature data frame and `st_transform()` to convert it from WGS 84 to Singapore SVY21 (EPSG:3414).

```{r}
#| label: import-mpsz
mpsz <- st_read(
  "data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson"
) %>%
  st_transform(crs = 3414)
```

#### Tidying data

The GeoJSON file already stores the required planning attributes in separate fields. Therefore, I retain the region, planning area, and subzone identifiers directly and keep the geometry column at the end.

```{r}
#| label: tidy-mpsz
mpsz <- mpsz %>%
  select(
    REGION_N,
    PLN_AREA_N,
    SUBZONE_N,
    SUBZONE_C,
    geometry
  )

mpsz
```

### Importing Attribute Data into R

I import the 2024 population CSV file as a tibble named `popdata`.

```{r}
#| label: import-population
popdata <- read_csv(
  "data/aspatial/respopagesextod2024.csv"
)
```

### Data Preparation

The population data is reorganised to derive the variables required for thematic mapping:

- **YOUNG:** residents aged 0–24;
- **ECONOMY ACTIVE:** residents aged 25–64;
- **AGED:** residents aged 65 and above;
- **TOTAL:** residents across all age groups; and
- **DEPENDENCY:** the ratio of young and aged residents to the economically active population.

#### Data wrangling

I first aggregate the population by planning area, subzone, and age group. I then use `pivot_wider()` to place each age group in a separate column before calculating the five required variables.

```{r}
#| label: wrangle-population
popdata2024 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(
    POP = sum(Pop),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = AG,
    values_from = POP
  ) %>%
  mutate(
    YOUNG = rowSums(
      pick(
        `0_to_4`,
        `5_to_9`,
        `10_to_14`,
        `15_to_19`,
        `20_to_24`
      )
    ),
    `ECONOMY ACTIVE` = rowSums(
      pick(
        `25_to_29`,
        `30_to_34`,
        `35_to_39`,
        `40_to_44`,
        `45_to_49`,
        `50_to_54`,
        `55_to_59`,
        `60_to_64`
      )
    ),
    AGED = rowSums(
      pick(
        `65_to_69`,
        `70_to_74`,
        `75_to_79`,
        `80_to_84`,
        `85_to_89`,
        `90_and_over`
      )
    ),
    TOTAL = YOUNG + `ECONOMY ACTIVE` + AGED,
    DEPENDENCY = (YOUNG + AGED) / `ECONOMY ACTIVE`
  ) %>%
  select(
    PA,
    SZ,
    YOUNG,
    `ECONOMY ACTIVE`,
    AGED,
    TOTAL,
    DEPENDENCY
  )
```

#### Joining the Attribute and Geospatial Data

The `PA` and `SZ` values contain both upper- and lowercase letters, while the corresponding boundary fields are uppercase. I convert the two population identifiers to uppercase and remove subzones with no economically active population.

```{r}
#| label: standardise-identifiers
popdata2024 <- popdata2024 %>%
  mutate_at(
    .vars = vars(PA, SZ),
    .funs = list(toupper)
  ) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

I use `left_join()` with the geospatial data on the left so that the result remains a simple feature data frame. `SUBZONE_N` and `SZ` are the common identifiers.

```{r}
#| label: join-population-boundaries
mpsz_pop2024 <- left_join(
  mpsz,
  popdata2024,
  by = c("SUBZONE_N" = "SZ")
)
```

I save the joined data as an RDS file for reuse.

```{r}
#| label: save-joined-data
write_rds(
  mpsz_pop2024,
  "data/rds/mpsz_pop2024.rds"
)
```

## Choropleth Mapping Geospatial Data Using **tmap**

Choropleth maps represent enumeration units with graduated colours. In this exercise, the enumeration units are planning subzones and the main mapped variable is the dependency ratio.

### Plotting a Choropleth Map Quickly Using `qtm()`

I set **tmap** to static plot mode and use `qtm()` to create a quick choropleth map of dependency ratio.

```{r}
#| label: quick-choropleth
#| fig-width: 8
#| fig-height: 6
tmap_mode("plot")

qtm(
  shp = mpsz_pop2024,
  fill = "DEPENDENCY"
)
```

### Creating a Choropleth Map Using **tmap** Elements

For greater control over individual map elements, I build the map using **tmap** layers.

```{r}
#| label: full-choropleth
#| fig-width: 9
#| fig-height: 7
tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(
      title = "Dependency ratio"
    )
  ) +
  tm_title(
    "Distribution of Dependency Ratio by Planning Subzone"
  ) +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2) +
  tm_credits(
    paste0(
      "Source: Planning subzone boundary from Urban Redevelopment ",
      "Authority (URA)\nand population data from Department of Statistics"
    ),
    position = c("left", "bottom")
  )
```

#### Drawing a Base Map

`tm_shape()` defines the input spatial data and `tm_polygons()` draws the planning subzone polygons.

```{r}
#| label: base-map
#| fig-width: 8
#| fig-height: 6
tm_shape(mpsz_pop2024) +
  tm_polygons()
```

#### Drawing a Choropleth Map Using `tm_polygons()`

Assigning `DEPENDENCY` to the fill aesthetic shades each planning subzone according to its dependency ratio.

```{r}
#| label: polygons-choropleth
#| fig-width: 8
#| fig-height: 6
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY")
```

#### Drawing a Choropleth Map Using `tm_fill()` and `tm_borders()`

`tm_polygons()` combines polygon fill and borders. The same components can be controlled separately with `tm_fill()` and `tm_borders()`.

```{r}
#| label: fill-only
#| fig-width: 8
#| fig-height: 6
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY")
```

```{r}
#| label: fill-and-borders
#| fig-width: 8
#| fig-height: 6
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders()
```

I then change the border colour, width, and line type.

```{r}
#| label: styled-borders
#| fig-width: 8
#| fig-height: 6
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders(
    col = "grey60",
    lwd = 0.1,
    lty = "dashed"
  )
```

### Data Classification Methods of **tmap**

Data classification groups observations into ranges or classes. **tmap** supports methods including fixed, standard deviation, equal interval, pretty, quantile, k-means, hierarchical clustering, bagged clustering, Fisher, and Jenks.

#### Plotting Choropleth Maps with Built-in Classification Methods

The first map uses five quantile classes.

```{r}
#| label: quantile-classification
#| fig-width: 8
#| fig-height: 6
tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5
    )
  ) +
  tm_borders(fill_alpha = 0.5)
```

The next map uses five equal-width classes.

```{r}
#| label: equal-classification
#| fig-width: 8
#| fig-height: 6
tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "equal",
      n = 5
    )
  ) +
  tm_borders(fill_alpha = 0.5)
```

The quantile method distributes the subzones more evenly among classes, whereas equal intervals retain equal value ranges.

#### Plotting a Choropleth Map with Custom Breaks

Before defining the breaks, I examine descriptive statistics for `DEPENDENCY`.

```{r}
#| label: dependency-summary
summary(mpsz_pop2024$DEPENDENCY)
```

I use six break points to form five classes.

```{r}
#| label: custom-breaks
#| fig-width: 8
#| fig-height: 6
tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      breaks = c(
        0,
        0.60,
        0.70,
        0.80,
        0.90,
        1.00
      )
    )
  ) +
  tm_borders(fill_alpha = 0.5)
```

### Colour Scheme

**tmap** supports colour ramps from **RColorBrewer** as well as user-defined palettes.

#### Using a ColourBrewer Palette

I apply a green sequential palette to the dependency ratio.

```{r}
#| label: green-palette
#| fig-width: 8
#| fig-height: 6
tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "brewer.greens"
    )
  ) +
  tm_borders(fill_alpha = 0.5)
```

Adding a minus sign reverses the colour sequence.

```{r}
#| label: reversed-green-palette
#| fig-width: 8
#| fig-height: 6
tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "-brewer.greens"
    )
  ) +
  tm_borders(fill_alpha = 0.5)
```

#### Cartographic Furniture

I add a compass, scale bar, grid lines, and data credits.

```{r}
#| label: cartographic-furniture
#| fig-width: 9
#| fig-height: 7
tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5
    )
  ) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits(
    "Source: data.gov.sg & SingStat",
    position = c("left", "bottom")
  )
```

### Map Layout

Map layout combines the frame, typography, scale, margins, legend, and other map elements.

#### Map Legend

I provide a clear legend title and allow **tmap** to place the legend inside the available map space.

```{r}
#| label: map-legend
#| fig-width: 9
#| fig-height: 7
tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5
    ),
    fill.legend = tm_legend(
      title = "Dependency ratio"
    )
  ) +
  tm_pos_auto_in() +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits(
    "Source: data.gov.sg & SingStat",
    position = c("left", "bottom")
  )
```

#### Map Style

I apply the natural style and the reversed green palette.

```{r}
#| label: natural-style
#| fig-width: 8
#| fig-height: 6
tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "-brewer.greens"
    )
  ) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")
```

I reset the default style to white before preparing the remaining maps.

```{r}
#| label: reset-style
#| include: false
tmap_style("white")
```

## Drawing Small Multiple Choropleth Maps

Small multiple maps place several related maps side by side so that their spatial patterns can be compared.

### Assigning Multiple Values to an Aesthetic Argument

Assigning `YOUNG` and `AGED` to the fill aesthetic creates one map for each variable.

```{r}
#| label: multiple-fill-values
#| fig-width: 10
#| fig-height: 6
tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = c("YOUNG", "AGED"),
    fill.legend = tm_legend(
      position = tm_pos_in("right", "bottom")
    ),
    fill.scale = tm_scale_intervals(
      style = "equal",
      n = 5,
      values = "brewer.blues"
    )
  ) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")
```

### Arranging Multiple Choropleth Maps in a Grid

I create separate young and aged population maps and arrange them in two columns.

```{r}
#| label: arranged-maps
#| fig-width: 11
#| fig-height: 6
youngmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "YOUNG",
    fill.legend = tm_legend(
      position = tm_pos_in("right", "bottom"),
      item.height = 0.8
    ),
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.blues"
    )
  ) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Young Population")

agedmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "AGED",
    fill.legend = tm_legend(
      position = tm_pos_in("right", "bottom"),
      item.height = 0.8
    ),
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.blues"
    )
  ) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Aged Population")

tmap_arrange(
  youngmap,
  agedmap,
  asp = 1,
  ncol = 2
)
```

### Defining a Group-by Variable in `tm_facets()`

I use `REGION_N` as the faceting variable so that the dependency ratio is shown separately for each region.

```{r}
#| label: region-facets
#| fig-width: 10
#| fig-height: 8
tm_shape(mpsz_pop2024) +
  tm_fill(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.blues"
    )
  ) +
  tm_facets(
    by = "REGION_N",
    nrow = 2,
    ncol = 3,
    free.coords = TRUE,
    drop.units = TRUE
  ) +
  tm_layout(
    legend.show = TRUE,
    title.position = c("center", "center"),
    title.size = 20
  ) +
  tm_borders(fill_alpha = 0.5)
```

## Mapping Spatial Objects Meeting a Selection Criterion

Instead of drawing all regions, I use `filter()` to select the Central Region and map its dependency ratio.

```{r}
#| label: central-region
#| fig-width: 8
#| fig-height: 7
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.greens"
    ),
    fill.legend = tm_legend()
  ) +
  tm_borders(fill_alpha = 0.5)
```

## Complementing a Thematic Map with a Statistical Chart

Maps show spatial patterns while statistical charts show the distribution of numerical values. I add a box plot to the Central Region map.

```{r}
#| label: map-with-boxplot
#| fig-width: 9
#| fig-height: 7
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.greens"
    ),
    fill.legend = tm_legend(),
    fill.chart = tm_chart_box()
  ) +
  tm_borders() +
  tm_layout(asp = 0.8)
```

I identify dependency-ratio outliers with `boxplot.stats()`, outline them in red, and label their subzone names.

```{r}
#| label: highlight-outliers
#| fig-width: 9
#| fig-height: 7
mpsz_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(
  mpsz_selected$DEPENDENCY
)

outlier_vals <- stats$out

outlier_sf <- mpsz_selected[
  mpsz_selected$DEPENDENCY %in% outlier_vals,
]

tm_shape(mpsz_selected) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(),
    fill.chart = tm_chart_box()
  ) +
  tm_borders(fill_alpha = 0.5) +
  tm_shape(outlier_sf) +
  tm_borders(
    col = "red",
    lwd = 2
  ) +
  tm_text(
    "SUBZONE_N",
    col = "red",
    size = 0.7
  ) +
  tm_layout(asp = 0.8)
```

## Creating an Interactive Map

I switch **tmap** to interactive view mode so that the Central Region map can be explored by panning, zooming, and selecting subzones.

```{r}
#| label: interactive-map
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(
  region_selected$DEPENDENCY
)

outlier_vals <- stats$out

outlier_sf <- region_selected[
  region_selected$DEPENDENCY %in% outlier_vals,
]

tmap_mode("view")

tm_shape(
  region_selected,
  bbox = region_bbox
) +
  tm_fill(
    "DEPENDENCY",
    id = "SUBZONE_N",
    popup.vars = c(
      "Name" = "SUBZONE_N",
      "Dependency" = "DEPENDENCY"
    )
  ) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(
    col = "red",
    lwd = 2
  )
```

The first interactive map allows unrestricted zooming. I use `set_zoom_limits` to keep the map within a practical zoom range.

```{r}
#| label: interactive-map-zoom-limits
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(
  region_selected$DEPENDENCY
)

outlier_vals <- stats$out

outlier_sf <- region_selected[
  region_selected$DEPENDENCY %in% outlier_vals,
]

tmap_mode("view")

tm_shape(
  region_selected,
  bbox = region_bbox
) +
  tm_fill(
    "DEPENDENCY",
    id = "SUBZONE_N",
    popup.vars = c(
      "Name" = "SUBZONE_N",
      "Dependency" = "DEPENDENCY"
    )
  ) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(
    col = "red",
    lwd = 2
  ) +
  tm_view(
    set_zoom_limits = c(12, 14)
  )
```

```{r}
#| label: restore-plot-mode
#| include: false
tmap_mode("plot")
```

## References

- [tmap: Thematic Maps in R](https://r-tmap.github.io/tmap/)
- [sf: Simple Features for R](https://r-spatial.github.io/sf/)
- [dplyr](https://dplyr.tidyverse.org/)
- [tidyr](https://tidyr.tidyverse.org/)
